---
title: "Getting started with actiRhythm"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Getting started with actiRhythm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 110
)
```

You have an ActiGraph recording for one person and a practical question. Does this
person have a stable daily rhythm, how strong is it, and is it real or could it be
noise? This article answers that question end to end on one bundled recording. By
the end you will be able to read the file, describe the rhythm without assuming a
shape, test whether it is real, locate and bound its period, probe its finer
structure, and write out a one-row summary you can stack across people.

The method assumes a multi-day activity-count series with timestamps, which is what
an ActiGraph `.agd` file holds and what counts exported from any source can supply.
Every number and figure below is reproducible from the installed package.

```{r libraries, message = FALSE, warning = FALSE}
library(actiRhythm)
library(ggplot2)
```

## The recording

`read.agd()` opens the `.agd` SQLite file and auto-detects the epoch length and
axes; `agd.counts()` returns a tidy data frame with `$axis1` (the vertical-axis
counts) and `$timestamp`. One recording ships with the package, reached through
`example_agd()`.

```{r read}
agd <- agd.counts(read.agd(example_agd(1), verbose = FALSE))
head(agd[, c("timestamp", "axis1")])
```

Read the actogram top to bottom before computing anything. It is the fastest way to
see whether there is a rhythm to measure at all.

```{r actogram, fig.height = 5, fig.cap = "Double-plotted actogram: each row is one day shown twice across 48 hours, so a stable rhythm forms a vertical band of activity at the same clock time each day; scattered fill marks a fragmented or shifting rhythm."}
plot_actogram(agd$axis1, agd$timestamp)
```

In this recording the active rows do not stack into one clean vertical band; they
slide a little later or earlier from day to day, which is the low interdaily
stability the next section puts a number on.

## The nonparametric rhythm

`circadian.rhythm()` describes the rest-activity pattern without assuming a
waveform. It returns interdaily stability (IS) and intradaily variability (IV),
introduced by Witting et al. (1990), together with the relative amplitude (RA) and
the least- and most-active windows L5 and M10 of van Someren et al. (1999), each
with its onset time.

```{r rhythm}
rhythm <- circadian.rhythm(agd$axis1, agd$timestamp)
rhythm
```

Read the three headline numbers together. IS ranges 0 to 1 and rises as the daily
pattern repeats more exactly from day to day; IV runs from about 0 (a smooth sine)
to about 2 (noisy or split into ultradian bouts); RA ranges 0 to 1 and grows with
the contrast between an active day and a restful night. Here a high RA (about 0.98)
sits with a low IS (about 0.23): a strong day-night contrast carried on irregular
timing. The header also notes that six of the eight calendar days qualified, because
`circadian.rhythm()` sets aside days with too little valid recording before
computing the metrics.

## Cosinor and the rhythmicity test

`cosinor.analysis()` fits a 24-hour cosine and returns the MESOR (rhythm-adjusted
mean), amplitude, and acrophase, the clock time of the peak (Cornelissen 2014).

```{r cosinor}
cos <- cosinor.analysis(agd$axis1, agd$timestamp, period = 24)
c(mesor = unname(cos$mesor), amplitude = unname(cos$amplitude),
  acrophase = unname(cos$acrophase))
```

The acrophase is **circular**: it is a clock time, so 23.5 and 0.5 are an hour
apart, not 23 hours apart, and it is only interpretable once the amplitude is
distinguishable from zero. `rhythmicity.test()` is that check, the zero-amplitude
F-test of Nelson et al. (1979), reported with the percent of variance the rhythm
explains.

```{r rhythmicity}
rhythmicity.test(agd$axis1, agd$timestamp, cosinor_result = cos)
```

A small p-value rejects the flat-line null. Here the test is clearly significant
(p well under 0.05) and the single cosine explains a little over forty percent of
the variance, so the acrophase above is worth reading. The percent-rhythm also
reminds you that more than half the variance is structure a single cosine misses.
Overlay the fit to see what it is and is not catching.

```{r cosinor-plot, fig.cap = "Hourly activity with the fitted 24-hour cosine overlaid; the peak of the curve is the acrophase and its height above the MESOR line is the amplitude. Systematic departures from the curve are where a single cosine is too smooth for real rest-activity data."}
plot_extended_cosinor(agd$axis1, agd$timestamp)
```

Look where the fitted curve and the hourly points part company: the single cosine
runs too smooth through the early hours, and that gap is the structure the
percent-rhythm says one cosine leaves on the table. When you need the joint
uncertainty in amplitude and acrophase together, `cosinor.confidence.ellipse(cos)`
returns the Bingham et al. (1982) ellipse; an ellipse that encloses the origin is
the geometric form of "no detectable rhythm".

## The period

A cosinor assumes exactly 24 hours. `circadian.period()` does not: it runs a
Lomb-Scargle periodogram over the 18 to 30 hour band and returns the dominant period
`tau`, with a false-alarm probability in `$p_value` (Lomb 1976; Scargle 1982;
Baluev 2008).

```{r period}
per <- circadian.period(agd$axis1, agd$timestamp)
c(tau = per$tau, p_value = per$p_value)
```

The strongest cycle here sits near 25.4 hours, meaningfully longer than a solar day,
and the false-alarm probability is essentially zero, so the cycle itself is not in
doubt. Its exact length is less certain. `period.ci()` refines that peak and
bootstraps it with a circular block bootstrap (Kunsch 1989; Politis and Romano
1992), which respects the autocorrelation in activity data, and the interval it
returns is wide enough to span 24 hours. On a single recording the period is clearly
present but not pinned to the hour. Passing `seed = 1` makes the bootstrap
reproducible; the 50 replicates here keep the vignette fast, and a real interval
would use more.

```{r period-ci}
period.ci(agd$axis1, agd$timestamp, n_boot = 50, seed = 1)
```

```{r periodogram, fig.cap = "Lomb-Scargle spectral power across candidate periods; the tallest peak is the dominant rhythm, and a peak rising above the significance line rejects the no-rhythm null."}
plot_periodogram(agd$axis1, agd$timestamp)
```

The chi-square periodogram of Sokolove and Bushell (1978) asks the same question a
different way, scoring each trial period by how much of the variance folds onto it.
A second estimator that clears its own threshold at the same period is reassurance
that the peak is real rather than an artifact of one method.

```{r chisq, fig.cap = "Chi-square (Sokolove-Bushell) periodogram: the Qp statistic against trial period, with the significance threshold. The peak should fall at the period the Lomb-Scargle periodogram found."}
plot_chisq(agd$axis1, agd$timestamp)
```

## Watching the period drift

A single periodogram averages over the whole recording. `circadian.spectrogram()`
slides a window across it and recomputes the periodogram in each, so you can see the
dominant period move, the chronobiology equivalent of a music spectrogram.

```{r spectrogram, fig.height = 4, fig.cap = "Period (vertical) against time (horizontal), coloured by spectral power. A horizontal band of colour is a stable period; a band that bends upward or downward is a rhythm lengthening or shortening across the recording."}
circadian.spectrogram(agd$axis1, agd$timestamp, step_hours = 24)$plot
```

Follow the band of colour left to right: here it drifts upward, the dominant period
lengthening across the recording rather than holding at 24 hours. Read the later
windows with some caution, since they ride the upper edge of the search band, where
a weakening rhythm and a genuinely longer period can look alike.

## Nonlinear structure

Two recordings can share the same IS and amplitude yet differ in their
moment-to-moment dynamics. `fractal.dfa()` estimates the detrended-fluctuation
exponent alpha, the long-range temporal correlation in the series (Peng et al. 1994;
Hu et al. 2001).

```{r dfa}
fractal.dfa(agd$axis1)$alpha
```

An alpha near 1 is the "1/f" scaling typical of healthy physiological output; near
0.5 is uncorrelated noise; well above 1 is a smoother, more random-walk-like signal.
Here alpha comes out near 0.93, close to the 1/f end, so this series carries genuine
long-range correlation rather than looking like noise. The log-log fit shows whether
one exponent holds across all time scales.

```{r dfa-plot, fig.cap = "Detrended fluctuation on log-log axes; the slope is the DFA exponent alpha. One straight line is monofractal scaling, a visible kink is a crossover between short- and long-range dynamics."}
plot_dfa(agd$axis1)
```

Two further questions use the same series and a single call each.
`multiscale.entropy()` asks how the complexity of the signal changes as you
coarse-grain it across time scales (Costa et al. 2002), and `mfdfa()` asks whether
one scaling exponent is enough or the series is genuinely multifractal (Kantelhardt
et al. 2002). Their help pages carry the math; here it is enough to know they refine,
not replace, the single alpha above.

## A fuller profile, and a decomposition

The cosinor fits one cosine to the day. `circadian.flm()` fits a functional model of
the same 24-hour profile with several Fourier harmonics, so it follows the real,
asymmetric shape of the day instead of forcing a symmetric curve (Wang et al. 2011);
the single cosine is its one-harmonic special case.

```{r flm}
flm <- circadian.flm(agd$axis1, agd$timestamp)
flm
```

```{r flm-plot, fig.cap = "The functional fit (line) over the averaged hourly profile (points). The extra Fourier harmonics let the curve follow the asymmetric morning rise and evening decline that a single cosine smooths over."}
ggplot(flm$smooth_curve, aes(t, activity)) +
  geom_point(data = flm$fitted_profile, aes(t, observed), colour = "grey55") +
  geom_line(colour = "#236192", linewidth = 1) +
  labs(x = "Hour of day", y = "Activity") +
  theme_actiRhythm()
```

The extra harmonics buy back the variance the single cosine left on the table, and
the curve now bends through the morning rise the one cosine ran straight past.

`circadian.ssa()` takes the opposite view. Instead of an average day it decomposes
the whole recording with singular spectrum analysis into additive components: a slow
trend, a circadian pair, and noise (Golyandina and Zhigljavsky 2013). The trajectory
matrix grows with the series, so bin a long minute recording to a coarser epoch
first.

```{r ssa}
bin      <- as.integer(as.numeric(agd$timestamp) %/% 600)   # 10-minute bins
counts10 <- tapply(agd$axis1, bin, sum)
t10      <- as.POSIXct(as.numeric(names(counts10)) * 600, origin = "1970-01-01", tz = "UTC")
ssa <- circadian.ssa(as.numeric(counts10), t10)
ssa
```

```{r ssa-plot, fig.height = 3.5, fig.cap = "The circadian component singular spectrum analysis pulls out of the whole recording: the daily rhythm separated from the slow trend and the noise."}
ggplot(data.frame(time = t10, circadian = ssa$circadian), aes(time, circadian)) +
  geom_line(colour = "#236192") +
  labs(x = "Time", y = "Circadian component") +
  theme_actiRhythm()
```

For this recording the circadian pair carries about 14 percent of the variance with a
fundamental period near 24.5 hours, the same slightly-long period the periodogram
found earlier, now isolated as a clean component of the raw series.

## Rest-activity transitions

`state.transitions()` summarises how readily the subject switches between rest and
activity, the kRA (rest-to-active) and kAR (active-to-rest) rates that capture
fragmentation a single amplitude cannot (Lim et al. 2011).

```{r transitions}
st <- state.transitions(agd$axis1)
c(kRA = st$kRA, kAR = st$kAR)
```

For this subject kAR runs well above kRA: once active they settle back to rest
quickly, but once at rest they stay there. That asymmetry is the fragmentation a
single amplitude cannot see.

## Pinpointing sleep and wake

`state.transitions()` gives one pair of rates for the whole recording.
`sleep.changepoints()` instead locates the sleep-onset and wake-onset of each night
directly from the counts, with no scored sleep required: a 24-hour cosinor bounds
each rest and active span roughly, and a change point inside each bound places the
precise transition (Chen and Sun 2024).

```{r changepoints}
cp <- sleep.changepoints(agd$axis1, agd$timestamp)
cp
```

```{r cp-plot, fig.height = 3.5, fig.cap = "Each detected sleep onset (blue) and wake onset (orange) by date and clock time. The detector reads the per-night sleep and wake timing directly from the counts."}
cps <- cp$changepoints
cps$clock <- as.numeric(format(cps$time, "%H")) + as.numeric(format(cps$time, "%M")) / 60
ggplot(cps, aes(time, clock, colour = type)) +
  geom_point(size = 3) +
  scale_colour_manual(values = c("sleep onset" = "#236192", "wake onset" = "#E69F00")) +
  labs(x = "Date", y = "Clock hour", colour = NULL) +
  theme_actiRhythm()
```

For this recording the detector finds seven nights, sleep onset late in the evening
and wake onset around seven in the morning, with most nights near eight to nine hours
of rest. That is the per-night sleep and wake timing the single transition rate above
cannot localise.

## Every rest bout, naps included

`sleep.changepoints()` finds the one main rest bout of each cycle. `rest.periods()`
takes the complementary view, consolidating every spell of low activity into a rest
bout, however many a day holds, daytime naps and fragmented rest included
(Roenneberg et al. 2015). Each epoch is compared to a fraction of its own 24-hour
activity level, and runs that stay below it grow into consolidated bouts.

```{r restperiods}
rp <- rest.periods(agd$axis1, agd$timestamp)
rp
```

```{r restperiods-plot, fig.height = 3, fig.cap = "Each consolidated rest bout as a bar from onset to offset. Detecting more than one bout a day, including naps, is what separates this from the one-per-cycle change-point detector."}
ggplot(rp$rest_periods, aes(onset, bout, xend = offset, yend = bout, colour = type)) +
  geom_segment(linewidth = 3) +
  scale_colour_manual(values = c(main = "#236192", nap = "#E69F00")) +
  labs(x = "Time", y = "Rest bout", colour = NULL) +
  theme_actiRhythm()
```

For this recording it consolidates nine rest bouts across six days, more than one a
day, so a few days carry a second rest beyond the main night.

`rest.crespo()` reaches a comparable result by a different route, rank-order
filtering and binary morphology rather than seed-and-grow consolidation (Crespo et
al. 2012); running both gives an independent cross-check of where a recording's rest
bouts fall. Neither applies a wear-time filter, so gate the counts on valid wear
first if a recording has device-off stretches.

## Scoring sleep and its regularity

The metrics so far describe the rhythm; the next need a per-epoch sleep or wake
label. `sleep.cole.kripke()` produces that label directly from the counts, the
standard count-based classifier for adults (Cole et al. 1992), with
`sleep.sadeh()` as the children-and-adolescents alternative.

```{r scoring}
state <- sleep.cole.kripke(agd$axis1)
table(state)
```

Count-based scoring labels every low-activity epoch as sleep, so a mostly sedentary
recording scores a high sleep fraction; what matters downstream is how regular that
label is from day to day. `sleep.regularity.index()` measures exactly that, the
probability that two epochs 24 hours apart share the same state (Phillips et al.
2017).

```{r sri}
sleep.regularity.index(state, agd$timestamp)
```

When you also have explicit sleep periods, `social.jet.lag()` returns the work-day
to free-day mid-sleep difference (Wittmann et al. 2006) and `lids()` the locomotor
inactivity during sleep that tracks ultradian sleep structure (Winnebeck et al.
2018).

## A wider set of methods

The core above answers most questions. actiRhythm also carries a wider set of
circadian methods for finer or more specialized work, shown here on the same
recording.

### Finer nonparametric metrics

`intradaily.variability.multiscale()` averages intradaily variability across bin
sizes, `activity.extrema()` generalizes L5 and M10 to any window with onset and
midpoint times, `dichotomy.index()` measures how cleanly rest separates from the
active day (Mormont et al. 2000), and `circadian.daily()` reports each day on its
own so within-recording drift shows. `rest.activity.fragmentation()` adds the
rest and active bout-length view that complements the kRA/kAR rates above.

```{r daily}
daily <- circadian.daily(agd$axis1, agd$timestamp)
dichotomy.index(agd$axis1, agd$timestamp, rest = state == "S")
```

```{r daily-plot, fig.height = 3, fig.cap = "The M10 onset (the start of the most active ten hours) for each day. Reporting it per day, rather than pooled, shows how steady the active phase is across the recording."}
ggplot(daily$daily, aes(date, M10_onset_h)) +
  geom_line(colour = "grey60") + geom_point(colour = "#236192", size = 2) +
  labs(x = NULL, y = "M10 onset (h)") +
  theme_actiRhythm()
```

### The shape and phase of the day

`cosinor.multicomponent()` adds harmonics to the single cosine and picks how many by
information criterion, `activity.onset.offset()` marks the daily activity onset and
offset by the relative-difference method (Roenneberg et al. 2003), and
`phase.concentration()` tests whether the daily phase markers cluster (Rayleigh and
Hermans-Rasson).

```{r shape}
cosinor.multicomponent(agd$axis1, agd$timestamp)
phase.concentration(daily$daily$M10_onset_h)
```

### Time-frequency and adaptive decomposition

On the ten-minute series from earlier, `circadian.wavelet()` gives the time-frequency
power spectrum, `ultradian.bandpower()` splits the variance into ultradian bands, and
`circadian.emd()` with `hilbert.huang()` extracts the circadian component
data-adaptively and tracks its instantaneous period.

```{r wavelet-plot, fig.height = 3, fig.cap = "The global wavelet power spectrum. The peak sits at the circadian period (dashed line at 24 hours), recovered without assuming a fixed cosine shape."}
wav <- circadian.wavelet(as.numeric(counts10), t10, epoch_length = 600)
ggplot(data.frame(period = wav$period_hours, power = wav$global_power),
       aes(period, power)) +
  geom_line(colour = "#236192") +
  geom_vline(xintercept = 24, linetype = 2, colour = "grey50") +
  scale_x_continuous(trans = "log2") +
  labs(x = "Period (h, log scale)", y = "Wavelet power") +
  theme_actiRhythm()
```

```{r emd}
emd <- circadian.emd(as.numeric(counts10), t10, epoch_length = 600)
hilbert.huang(emd)
```

### Registration, residual structure, and states

`curve.registration()` aligns the days on their active-phase landmark and reports a
scale-invariant chronotype phase, `residual.spectrum()` examines what the cosinor
leaves behind, and `rest.hmm()` fits a state-space model whose decoded path gives the
probability of rest across the day.

```{r hmm-plot, fig.height = 3, fig.cap = "The 24-hour rest-probability profile from the state-space model: the share of each clock hour the decoded path spends in the rest state."}
hmm <- rest.hmm(as.numeric(counts10), t10)
ggplot(hmm$tod_profile, aes(hour, p_rest)) +
  geom_col(fill = "#236192") +
  labs(x = "Hour of day", y = "P(rest)") +
  theme_actiRhythm()
```

## From raw acceleration

Everything so far runs on activity counts. Raw accelerometer files also record the
gravity component, and with it body posture, which counts cannot carry. actiRhythm reads
raw ActiGraph `.gt3x`, Axivity `.cwa`, and GENEActiv `.bin` files, auto-calibrates them,
and derives the three raw metrics used across the field: ENMO, MAD, and the z-angle.

`raw.metrics()` takes a file path. Raw acceleration is far too large to bundle, so for
a reproducible illustration `example_raw()` synthesises a recording; a real file is read
the same way, with `raw.metrics("recording.cwa")`.

```{r raw-metrics}
raw <- example_raw(days = 2)        # synthetic 2-day recording (or a real file path)
m <- raw.metrics(raw, epoch = 60)   # per-epoch ENMO (mg), MAD, and the z-angle
head(m)
```

ENMO is the activity signal, and every method above takes it directly: pass it to
`circadian.rhythm()`, or let `circadian.raw()` run the whole analysis from the file.

```{r circadian-raw}
cr <- circadian.rhythm(m$ENMO, m$time)
c(IS = cr$IS, IV = cr$IV, RA = cr$RA)
```

Calibration is applied first inside `raw.metrics()`. `auto.calibrate()` finds the
per-axis gain and offset that return still periods to the 1 g sphere (van Hees 2014);
on data with a known distortion it returns the original gain and offset to within a
fraction of a percent.

```{r calibrate}
set.seed(1)
u <- matrix(rnorm(40 * 3), 40, 3); u <- u / sqrt(rowSums(u^2))
cal <- do.call(rbind, lapply(seq_len(40), function(i)
  matrix(rep(u[i, ] / c(1.03, 0.97, 1.01) + c(0.04, -0.03, 0.02), each = 300),
         300, 3) + rnorm(900, 0, 0.004)))
auto.calibrate(data.frame(x = cal[, 1], y = cal[, 2], z = cal[, 3]),
               fs = 30)[c("scale", "offset")]
```

The z-angle supports a sleep detector that needs no diary and works from posture,
which the counts cannot do. `rest.spt()` finds the nightly sleep-period-time window from the
distribution of angle change (HDCZA, van Hees 2018), `sib.vanhees()` scores
sustained-inactivity bouts (van Hees 2015), and `sleep.from.spt()` combines them into
onset, wake and efficiency. `detect.nonwear.raw()` flags a stationary, taken-off
device (low standard deviation and range over the hour, van Hees 2011), and passing
its mask as `wear` keeps a device-off stretch from being read as one long night.
Computed here on the synthetic recording at a 5-second epoch, with the detected
windows shaded.

```{r spt, fig.height = 3, fig.cap = "The z-angle of the synthetic recording at 5-second epochs, with the two nightly sleep-period-time windows detected from the angle alone (HDCZA), gated by raw non-wear, shaded."}
m5    <- raw.metrics(raw, epoch = 5, metrics = "anglez")
wear  <- detect.nonwear.raw(raw, epoch = 5)
spt   <- rest.spt(m5$anglez, m5$time, epoch_length = 5, wear = wear)
sib   <- sib.vanhees(m5$anglez, epoch_length = 5)
sleep <- sleep.from.spt(spt, sib, m5$time, epoch_length = 5)
thin <- seq(1, nrow(m5), by = 12)
ggplot() +
  geom_rect(data = spt, aes(xmin = onset, xmax = offset, ymin = -90, ymax = 90),
            fill = "#236192", alpha = 0.18) +
  geom_line(data = m5[thin, ], aes(time, anglez), linewidth = 0.2, colour = "grey30") +
  labs(x = NULL, y = "z-angle (deg)") +
  theme_actiRhythm()
sleep[, c("date", "onset", "offset", "tst", "efficiency")]
```

Two more estimators measure fragmentation, and both run on the counts already loaded:
`activity.balance.index()` maps a detrended-fluctuation exponent to a 0 to 1 score that
peaks at the healthy 1/f balance (Danilevicz 2024), and `transition.probability()` gives
the closed-form rest-to-active and active-to-rest transition probabilities.

```{r abi-tp}
activity.balance.index(fractal.dfa(agd$axis1))
transition.probability(agd$axis1)[c("tp_ra_mle", "tp_ar_mle")]
```

### Agreement with GGIR

The raw metrics and the z-angle sleep detector reimplement the van Hees algorithms
that GGIR established (Migueles et al. 2019), so the whole chain should reproduce
GGIR's output, not just be internally consistent: calibration, the raw metrics, and
the z-angle sleep detection. Run side by side on the same real 7-day wrist recording
(118,800 five-second epochs), they agree at every stage.

Calibration matches to about three decimals, with scale 1.001 / 0.995 / 0.999 and the
same 0.006 g calibration error, and the offsets agree in magnitude once the two sign
conventions are lined up: GGIR applies `scale * x + offset`, actiRhythm `(x - offset) * scale`.
Per-epoch ENMO and z-angle both correlate at r = 0.99, with mean absolute differences
of 3.6 mg and 2.2 degrees. The HDCZA sleep-period window lands within about five to ten
seconds of GGIR each night, and total sleep time and WASO agree to about a minute on the
valid nights. Non-wear matches to within 0.05 per worn day, and the wider gap between
0.41 and 0.52 is entirely the parked-device tail that both tools already exclude.

The comparison is reproducible on your own file with GGIR installed:

```{r ggir-validation, eval = FALSE}
library(GGIR)
P <- load_params(); P$params_general$windowsizes <- c(5, 900, 3600)
I <- g.inspectfile(file, params_rawdata = P$params_rawdata, params_general = P$params_general)
ggir <- g.getmeta(file, params_rawdata = P$params_rawdata, params_general = P$params_general,
                  params_cleaning = P$params_cleaning, inspectfileobject = I)$metashort
acti <- raw.metrics(file, epoch = 5)

# Align on wall-clock time: the .gt3x stores local time and the two tools anchor it
# slightly differently, so match the clock label rather than absolute seconds.
gkey <- gsub("T", " ", substr(as.character(ggir$timestamp), 1, 19))
akey <- format(acti$time, "%Y-%m-%d %H:%M:%S")
m <- merge(data.frame(k = gkey, ge = ggir$ENMO * 1000, ga = ggir$anglez),
           data.frame(k = akey, ae = acti$ENMO,        aa = acti$anglez), by = "k")
c(ENMO = cor(m$ge, m$ae), anglez = cor(m$ga, m$aa))
```

The sleep figures come from `rest.spt()` and `sleep.from.spt()` run against GGIR's full
part 1 to 4 pipeline (`HASPT.algo = "HDCZA"`). On every clean night the window and the
parameters match to seconds and to about a minute. They diverge only on nights GGIR
flags as heavily invalid, 40 to 53% non-wear, where the two handle wear differently.
That is an edge case on badly degraded data, not a difference in the algorithm.

ActiGraph idle-sleep gaps are imputed as zeros by `read.gt3x`, which would otherwise
collapse the z-angle to a constant during quiescent periods. Like GGIR, actiRhythm
carries the last gravity vector forward through those gaps, so the angle stays correct
through the still periods the sleep detector relies on.

## Pooling the evidence

Each test above answers "is there a rhythm?" in its own language.
`consensus.rhythmicity()` runs four of them, the cosinor F-test, the Bingham
ellipse, the Lomb-Scargle false-alarm probability, and the chi-square periodogram,
and reports both a majority vote and a Fisher-combined p-value (Fisher 1925).

```{r consensus}
consensus.rhythmicity(agd$axis1, agd$timestamp)
```

For this recording three of the four tests agree that a rhythm is present and the
Fisher-combined p-value is effectively zero, so the verdict is unambiguous even
where one method on its own might hesitate. A one-row summary is then easy to
assemble and `rbind()` across subjects.

```{r summary}
data.frame(
  IS        = rhythm$IS,
  IV        = rhythm$IV,
  RA        = rhythm$RA,
  tau       = per$tau,
  amplitude = cos$amplitude,
  acrophase = cos$acrophase,
  dfa_alpha = fractal.dfa(agd$axis1)$alpha
)
```

Read across, this row is one person: a strongly amplitude-modulated rhythm whose
timing is irregular from day to day, running on a period a little over 24 hours.

## A whole study at once

You have now run the whole arc on one recording: read it, describe the rhythm
without assuming a shape, test whether it is real, locate and bound its period,
probe its finer structure, and pool the separate tests into one verdict, ending
with a single summary row. Most studies, though, are a folder of recordings rather
than one, and that is what `circadian.batch()` is for. It runs this entire arc over
every `.agd` file in a directory and returns one row per recording, so a cohort
becomes a single data frame you can model. A file that fails to read is reported in
an `error` column instead of stopping the run. The package ships two recordings, so
this runs as a real, if small, batch.

```{r batch}
batch <- circadian.batch(
  system.file("extdata", package = "actiRhythm"),
  verbose = FALSE
)
batch[, c("file", "IS", "IV", "RA", "period_tau", "rhythm_p_value")]
```

Each row is one subject, described by the same metrics as the walkthrough above, and
already the two differ: the second is steadier from day to day (higher IS) but less
amplitude-modulated (lower RA), and its rhythm clears significance only at the
margin. Point the call at your own folder and nothing else changes.

When you want the full analysis for a single recording rather than a summary row,
`circadian.workbook()` writes every table to a multi-sheet Excel file: a summary row
plus the hourly profile, both periodograms, the fluctuation and multifractal
spectra, and a data dictionary.

```{r workbook, eval = FALSE}
circadian.workbook(agd$axis1, agd$timestamp, file = "subject01.xlsx")
```

## Where next

- `?circadian.rhythm`, `?cosinor.analysis`, `?circadian.period`, `?fractal.dfa`,
  `?consensus.rhythmicity` give the per-function help with full references.
- `?circadian.batch` and `?circadian.workbook` are the study-scale entry points.
- For a multi-subject design, `population.cosinor()` and `cosinor.compare()` give
  the mean rhythm and a two-group test (Bingham/Hotelling); for covariates,
  repeated measures, or nesting, feed the per-subject metrics this package emits
  (MESOR, amplitude, acrophase, IS, IV, RA, period) into a mixed model with
  `lme4`/`nlme` alongside your design. The package gives you the metrics; you supply
  the multilevel model.
- Cite the package with `citation("actiRhythm")`, and report
  `packageVersion("actiRhythm")` with your results so the analysis stays
  reproducible across releases.
- Questions and bugs: <https://github.com/rdazadda/actiRhythm/issues>.

## References

Witting W, Kwa IH, Eikelenboom P, Mirmiran M, Swaab DF (1990). Alterations in the
circadian rest-activity rhythm in aging and Alzheimer's disease. Biological
Psychiatry 27(6):563-572. doi:10.1016/0006-3223(90)90523-5

Van Someren EJW, Swaab DF, Colenda CC, Cohen W, McCall WV, Rosenquist PB (1999).
Bright light therapy: improved sensitivity to its effects on rest-activity rhythms
in Alzheimer patients by application of nonparametric methods. Chronobiology
International 16(4):505-518. doi:10.3109/07420529908998724

Nelson W, Tong YL, Lee JK, Halberg F (1979). Methods for cosinor-rhythmometry.
Chronobiologia 6(4):305-323.

Cornelissen G (2014). Cosinor-based rhythmometry. Theoretical Biology and Medical
Modelling 11:16. doi:10.1186/1742-4682-11-16

Bingham C, Arbogast B, Cornelissen Guillaume G, Lee JK, Halberg F (1982). Inferential
statistical methods for estimating and comparing cosinor parameters. Chronobiologia
9(4):397-439.

Lomb NR (1976). Least-squares frequency analysis of unequally spaced data.
Astrophysics and Space Science 39(2):447-462. doi:10.1007/BF00648343

Scargle JD (1982). Studies in astronomical time series analysis. II. Statistical
aspects of spectral analysis of unevenly spaced data. The Astrophysical Journal
263:835-853. doi:10.1086/160554

Baluev RV (2008). Assessing the statistical significance of periodogram peaks.
Monthly Notices of the Royal Astronomical Society 385(3):1279-1285.
doi:10.1111/j.1365-2966.2008.12689.x

Kunsch HR (1989). The jackknife and the bootstrap for general stationary
observations. The Annals of Statistics 17(3):1217-1241. doi:10.1214/aos/1176347265

Politis DN, Romano JP (1992). A circular block-resampling procedure for stationary
data. In: LePage R, Billard L (eds), Exploring the Limits of Bootstrap, 263-270.
Wiley, New York.

Sokolove PG, Bushell WN (1978). The chi square periodogram: its utility for analysis
of circadian rhythms. Journal of Theoretical Biology 72(1):131-160.
doi:10.1016/0022-5193(78)90022-X

Peng C-K, Buldyrev SV, Havlin S, Simons M, Stanley HE, Goldberger AL (1994). Mosaic
organization of DNA nucleotides. Physical Review E 49(2):1685-1689.
doi:10.1103/PhysRevE.49.1685

Hu K, Ivanov PC, Chen Z, Carpena P, Stanley HE (2001). Effect of trends on detrended
fluctuation analysis. Physical Review E 64(1):011114. doi:10.1103/PhysRevE.64.011114

Costa M, Goldberger AL, Peng C-K (2002). Multiscale entropy analysis of complex
physiologic time series. Physical Review Letters 89(6):068102.
doi:10.1103/PhysRevLett.89.068102

Kantelhardt JW, Zschiegner SA, Koscielny-Bunde E, Havlin S, Bunde A, Stanley HE
(2002). Multifractal detrended fluctuation analysis of nonstationary time series.
Physica A 316(1-4):87-114. doi:10.1016/S0378-4371(02)01383-3

Lim ASP, Yu L, Costa MD, Buchman AS, Bennett DA, Leurgans SE, Saper CB (2011).
Quantification of the fragmentation of rest-activity patterns in elderly individuals
using a state transition analysis. Sleep 34(11):1569-1581. doi:10.5665/sleep.1400

Phillips AJK, Clerx WM, O'Brien CS, Sano A, Barger LK, Picard RW, Lockley SW, Klerman
EB, Czeisler CA (2017). Irregular sleep/wake patterns are associated with poorer
academic performance and delayed circadian and sleep/wake timing. Scientific Reports
7:3216. doi:10.1038/s41598-017-03171-4

Wittmann M, Dinich J, Merrow M, Roenneberg T (2006). Social jetlag: misalignment of
biological and social time. Chronobiology International 23(1-2):497-509.
doi:10.1080/07420520500545979

Winnebeck EC, Fischer D, Leise T, Roenneberg T (2018). Dynamics and ultradian
structure of human sleep in real life. Current Biology 28(1):49-59.e5.
doi:10.1016/j.cub.2017.11.063

Fisher RA (1925). Statistical Methods for Research Workers. Oliver and Boyd,
Edinburgh.

van Hees VT, Renstrom F, Wright A, et al. (2011). Estimation of daily energy
expenditure in pregnant and non-pregnant women using a wrist-worn tri-axial
accelerometer. PLoS ONE 6(7):e22922. doi:10.1371/journal.pone.0022922

van Hees VT, Gorzelniak L, Dean Leon EC, et al. (2013). Separating movement and
gravity components in an acceleration signal and implications for the assessment of
human daily physical activity. PLoS ONE 8(4):e61691. doi:10.1371/journal.pone.0061691

van Hees VT, Fang Z, Langford J, et al. (2014). Autocalibration of accelerometer data
for free-living physical activity assessment using local gravity and temperature.
Journal of Applied Physiology 117(7):738-744. doi:10.1152/japplphysiol.00421.2014

Vaha-Ypya H, Vasankari T, Husu P, Suni J, Sievanen H (2015). A universal, accurate
intensity-based classification of different physical activities using raw data of
accelerometer. Clinical Physiology and Functional Imaging 35(1):64-70.
doi:10.1111/cpf.12127

van Hees VT, Sabia S, Anderson KN, et al. (2015). A novel, open access method to
assess sleep duration using a wrist-worn accelerometer. PLoS ONE 10(11):e0142533.
doi:10.1371/journal.pone.0142533

van Hees VT, Sabia S, Jones SE, et al. (2018). Estimating sleep parameters using an
accelerometer without sleep diary. Scientific Reports 8:12975.
doi:10.1038/s41598-018-31266-z

Migueles JH, Rowlands AV, Huber F, Sabia S, van Hees VT (2019). GGIR: a research
community-driven open source R package for generating physical activity and sleep
outcomes from multi-day raw accelerometer data. Journal for the Measurement of
Physical Behaviour 2(3):188-196. doi:10.1123/jmpb.2018-0063

Danilevicz IM, van Hees VT, van der Heide FCT, et al. (2024). Measures of
fragmentation of rest activity patterns: mathematical properties and
interpretability. BMC Medical Research Methodology 24:132.
doi:10.1186/s12874-024-02255-w
