---
title: "Output codebook (metrics data dictionary)"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Output codebook (metrics data dictionary)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: REFERENCES.bib
csl: apa.csl
link-citations: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This is the lookup for every metric `actiRhythm` emits in its one-row summary:
the columns of `circadian.batch()`, the **Summary** sheet of
`circadian.workbook()`, and the slots of the individual result objects. Find any
output here: what it is, its formula or units, the range you should expect, the
defining reference, and where it lives in the result object.

The table below, the workbook's *Data Dictionary* sheet, and the package's tests
all read one file (`inst/extdata/metrics_dictionary.csv`), so the codebook can
never drift from the code (a test checks that the documented metrics are exactly
those `circadian.batch()` returns). Individual result objects carry further fields
beyond this summary; see each function's help page (for example
`?circadian.rhythm`) for the complete object.

```{r load, echo = FALSE}
dict <- read.csv(
  system.file("extdata", "metrics_dictionary.csv", package = "actiRhythm"),
  stringsAsFactors = FALSE
)
# Render identifiers and object slots as code (monospace, literal), and the
# free-text columns with their *, _, < and > shown literally rather than read as
# Markdown emphasis, math, or tags.
.code <- function(x) ifelse(nzchar(x), paste0("`", x, "`"), "")
.txt  <- function(x) {
  x <- gsub("([*_])", "\\\\\\1", x)
  x <- gsub("<", "&lt;", x, fixed = TRUE)
  gsub(">", "&gt;", x, fixed = TRUE)
}
show_family <- function(fam) {
  d <- dict[dict$family == fam, , drop = FALSE]
  knitr::kable(
    data.frame(
      Metric                   = .code(d$metric),
      Definition               = .txt(d$definition),
      `Formula / units`        = .txt(d$formula_units),
      `Range / interpretation` = .txt(d$range_interpretation),
      Reference                = d$reference,
      `Output object`          = .code(d$output_object),
      check.names = FALSE, stringsAsFactors = FALSE
    ),
    escape = FALSE, row.names = FALSE
  )
}
```

## Reading this codebook

Each metric is named exactly as it appears as a column of the `circadian.batch()`
data frame and the workbook **Summary** sheet. The **Output object** column shows
where the same value lives when you call the producing function directly:
`circadian.rhythm()$IS`, `cosinor.analysis()$mesor`, and so on. The **Reference**
column gives the citation key for the method; full references are listed at the
end. A metric whose **Range** notes "requires `sleep_state`" or "requires
`sleep_periods`" is `NA` unless that input is supplied.

## Recording descriptors

```{r io, echo = FALSE}
show_family("io")
```

## Nonparametric metrics

Shape-free descriptors of the rest-activity rhythm, computed with no assumed
waveform [@witting1990; @vansomeren1999].

```{r nonparametric, echo = FALSE}
show_family("nonparametric")
```

## Cosinor and rhythmicity

The single-component cosinor and its zero-amplitude rhythmicity test
[@cornelissen2014; @nelson1979].

```{r cosinor, echo = FALSE}
show_family("cosinor")
```

## Period and spectral

The endogenous period from the Lomb-Scargle and chi-square periodograms, with a
bootstrap confidence interval [@lomb1976; @sokolove1978].

```{r period, echo = FALSE}
show_family("period")
```

## Fractal and nonlinear

Long-range correlation and complexity of the activity series
[@peng1994; @kantelhardt2002; @costa2002].

```{r fractal, echo = FALSE}
show_family("fractal")
```

## Rest-activity transitions

State-transition rates and probabilities that capture fragmentation a single
amplitude cannot [@lim2011].

```{r phase, echo = FALSE}
show_family("phase")
```

## Sleep timing and regularity

Metrics that need a sleep label or sleep periods: the Sleep Regularity Index,
social jet lag, and LIDS [@phillips2017; @wittmann2006; @winnebeck2018].

```{r sleep, echo = FALSE}
show_family("sleep")
```

## References

Full citations for the method keys in the table above:
[@vansomeren1999; @witting1990; @cornelissen2014; @nelson1979; @lomb1976;
@baluev2008; @politis1992; @sokolove1978; @peng1994; @kantelhardt2002;
@costa2002; @lim2011; @phillips2017; @wittmann2006; @winnebeck2018].
