---
title: "Batch processing and reporting"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
bibliography: ../REFERENCES.bib
csl: ../apa.csl
link-citations: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, dpi = 110)
```

```{r libs, message = FALSE, warning = FALSE}
library(actiRhythm)
library(ggplot2)
```

## The idea, and a rule to remember

Every other article runs one method on one recording. A study runs every method on
every recording, and then has to put the numbers somewhere a collaborator can open.
That is what this family is for: `circadian.batch()` runs the whole analysis over a
folder of `.agd` files and returns **one tidy row per file**; `circadian.workbook()`
writes the complete analysis of a single recording to a multi-sheet Excel file; and
`circadian.daily()` breaks a recording into per-day rows so within-recording drift
shows instead of being averaged away. Above the single subject, `population.cosinor()`
and `cosinor.compare()` pool per-subject fits into a group rhythm and test whether two
groups differ.

The rule to carry through this article: **one row per file, and a bad file costs you a
cell, not the run.** `circadian.batch()` wraps each file in its own error handler, so a
recording that fails to read or analyse lands in an `error` column with `NA` metrics
while every other file completes. A batch that aborts on file 37 of 200 is worse than
useless; a batch that flags file 37 and finishes is a study you can act on.

## The shape of a run

There is no equation here (the family is plumbing, not a model), but the output has
a fixed shape worth stating up front.

- **`circadian.batch()`** returns a data frame: a `file` column, an `error` column,
  and then the full single-row summary that `circadian.workbook()` builds internally
  (the nonparametric IS/IV/RA/L5/M10, the cosinor MESOR/amplitude/acrophase and its
  rhythmicity F-test, the Lomb-Scargle and chi-square periods, DFA and multifractal
  width, multiscale entropy, and the rest-activity transition rates). Every file is
  one row; every metric is one column.
- **`circadian.workbook()`** writes that same summary as the first sheet of an
  `.xlsx`, then adds detail sheets (the hourly profile, both periodograms, the
  fluctuation and multifractal spectra, the transition curves) and a self-documenting
  **Data Dictionary** sheet, the same single source the
  [output codebook](output-codebook.html) renders from.
- **`circadian.daily()`** returns a `daily` data frame with one row per calendar day
  (`L5`, `M10`, their onset hours, `RA`, `IV`, and the day's total counts).

Knowing the shape is most of knowing how to script against it.

## Assumptions, and when they break

- **Readable, regularly-epoched files.** The batch assumes each path is an `.agd`
  ActiGraph file `read.agd()` can open. Anything else becomes an `error` row, which
  is the design, not a failure.
- **A consistent epoch across files.** By default the epoch is inferred per file from
  the timestamps; mixing 30 s and 60 s recordings is allowed, but the epoch-sensitive
  metrics (IV especially) are then only comparable within an epoch. Pass an explicit
  `epoch_length` to force one.
- **Enough valid days per file.** The per-day and between-day metrics need whole days;
  `circadian.daily()` drops any day with fewer than `min_hours` of valid recording
  before reporting it.
- **Subjects, not epochs, are the unit for the group tools.** `population.cosinor()`
  and `cosinor.compare()` fit one cosinor per subject and then pool; they assume each
  subject contributes an averaged 24-hour profile with at least `min_valid_hours` of
  data. They are not a substitute for a mixed model. See
  [Choosing a method](choosing-a-method.html) for when to reach for one.

## Recovering known truth

Before trusting the group tools on real cohorts, watch them recover a difference we
plant. We build two groups of six subjects each: a **morning** group whose rhythm peaks
at 08:00 with amplitude 45, and an **evening** group peaking at 13:00 with amplitude 38.
Both share a MESOR of 100. The planted between-group acrophase difference is exactly
five hours.

```{r known-truth}
set.seed(42)
hrs <- 0:23
act <- ts <- subj <- grp <- NULL
for (g in c("morning", "evening")) for (i in 1:6) {
  acro <- if (g == "morning") 8 else 13
  amp  <- if (g == "morning") 45 else 38
  y <- 100 + amp * cos(2 * pi * (hrs - acro) / 24) + rnorm(24, 0, 5)
  act  <- c(act, y)
  subj <- c(subj, rep(paste0(substr(g, 1, 1), i), 24))
  grp  <- c(grp, rep(g, 24))
  ts   <- c(ts, as.POSIXct("2024-01-01", tz = "UTC") + hrs * 3600)
}
ts <- as.POSIXct(ts, tz = "UTC", origin = "1970-01-01")
```

`population.cosinor()` pools each group's per-subject fits into a Bingham group-mean
rhythm with confidence intervals [@bingham1982]. It should return the planted MESOR,
amplitude, and acrophase for each group.

```{r pop-recover}
pc <- population.cosinor(act, ts, subject = subj, group = grp)
knitr::kable(
  data.frame(
    group     = c("morning", "evening"),
    planted_acrophase = c(8, 13),
    recovered_acrophase = c(pc$morning$acrophase, pc$evening$acrophase),
    planted_amplitude = c(45, 38),
    recovered_amplitude = c(pc$morning$amplitude, pc$evening$amplitude)
  ),
  digits = 2,
  caption = "The population cosinor recovers each group's planted acrophase and amplitude."
)
```

`cosinor.compare()` then tests whether the two groups differ. The omnibus is a Bingham
/ Hotelling \(T^2\) on the joint (MESOR, cosine, sine) vector [@bingham1982]; the
per-parameter rows break it down, with the acrophase test made circular-aware.

```{r compare-recover}
cmp <- cosinor.compare(act, ts, subject = subj, group = grp)
cmp
```

The omnibus is overwhelmingly significant, and the per-parameter breakdown lands where
we planted it: the MESOR difference is near zero, the amplitude differs by about the
planted seven units, and the **acrophase difference recovers the planted five hours**.
The tools report exactly the contrast we put in.

## On a real recording

The bundled `extdata` folder holds the package's example `.agd` files. Pointing
`circadian.batch()` at the directory analyses every file in it and returns one row each.

```{r batch}
folder <- system.file("extdata", package = "actiRhythm")
batch  <- circadian.batch(folder, verbose = FALSE)
nrow(batch)
batch[, c("file", "error", "IS", "IV", "RA",
          "cosinor_amplitude", "rhythm_p_value", "period_tau")]
```

Two files in, two rows out, with `error` all `NA` (both read and analysed cleanly),
and the full metric set sitting alongside in columns we have only sampled here. Each
row carries everything the [output codebook](output-codebook.html) documents.

To see *within* a recording rather than across files, `circadian.daily()` reports each
calendar day on its own.

```{r daily}
agd   <- agd.counts(read.agd(example_agd(1), verbose = FALSE))
daily <- circadian.daily(agd$axis1, agd$timestamp)
daily
```

```{r daily-plot, fig.height = 3, fig.cap = "The most-active-window onset for each day. Reporting it per day, rather than pooled into the batch row, shows how steady the active phase is across the recording."}
ggplot(daily$daily, aes(date, M10_onset_h)) +
  geom_line(colour = "grey60") +
  geom_point(colour = "#236192", size = 2) +
  labs(x = NULL, y = "M10 onset (h)") +
  theme_actiRhythm()
```

## Reading the numbers

The batch row is only as useful as your sense of what each column should be. A few
human yardsticks for scanning a study-sized table:

- **`error`** is the first column to read, not the last. Sort or filter on it before
  trusting any metric; a non-`NA` value means that row's numbers are missing, not that
  the subject has a flat rhythm.
- **`rhythm_p_value`** below 0.05 says the day-night cosine is real, but read it with
  `percent_rhythm` (how much of the day the single cosine captures) and `IS`/`IV` (how
  stable and how fragmented), exactly as in the [cosinor](cosinor.html) and
  [nonparametric](nonparametric.html) articles. The batch just lines those columns up
  side by side.
- **`period_tau`** near 24 is an entrained rhythm; a value drifting from 24 across a
  cohort is the cue to look at free-running, not a rounding error.
- For the **group comparison**, an omnibus \(T^2\) p-value answers "do the rhythms
  differ at all?" before you read any single parameter; the per-parameter rows then say
  *which* parameter carries the difference, with the acrophase row already unwrapped
  around the clock.

## The wider reporting family

The same machinery feeds two output formats a collaborator can open directly.

**A full single-subject workbook.** `circadian.workbook()` runs every method on one
recording and writes a multi-sheet `.xlsx`: the one-row summary, a detail sheet per
analysis, and a Data Dictionary. Here we write it to a temporary file and list its
sheets. (The chunk is guarded so it only runs where `openxlsx` is installed.)

```{r workbook, eval = requireNamespace("openxlsx", quietly = TRUE)}
xlsx <- tempfile(fileext = ".xlsx")
circadian.workbook(agd$axis1, agd$timestamp, file = xlsx, include_period_ci = FALSE)
openxlsx::getSheetNames(xlsx)
```

`circadian.batch()` accepts the same `file =` argument to write a single workbook whose
Summary sheet has one row per file, the study-level companion to the per-subject
workbook above.

**Saving a figure.** `save.circadian.plot()` is a thin wrapper over `ggplot2::ggsave()`
that takes any of the package's `plot_*` figures and writes it in the format implied by
the file extension, so a report script can emit its actograms without restating the
size and DPI each time.

```{r save-plot}
p   <- plot_actogram(agd$axis1, agd$timestamp)
png <- tempfile(fileext = ".png")
save.circadian.plot(p, png, width = 8, height = 5)
file.exists(png)
```

## Limitations

- **Speed scales with the slowest method, times the number of files.** The bootstrap
  period CI is off by default in `circadian.batch()` (`include_period_ci = FALSE`) for
  exactly this reason; turn it on only when you need it.
- **An `error` row is a missing subject, not a fixed one.** The batch keeps running, but
  you still have to go back and find out why a file failed to read.
- **The group tools pool, they do not model.** `population.cosinor()` and
  `cosinor.compare()` average per-subject fits and run a two-sample test; they do not
  borrow strength across subjects or handle nested designs. For repeated measures or
  covariates, the mixed-model route in [Choosing a method](choosing-a-method.html) is
  the right tool.
- **Column meanings live in one place.** Rather than memorise 40-plus column names,
  read them from the Data Dictionary sheet or the
  [output codebook](output-codebook.html); both render from the same source file.

## Reference and validation

The two group tools implement the population-mean cosinor and between-group comparison
of @bingham1982: `population.cosinor()` averages the linearized
per-subject coefficients into a Bingham group rhythm with confidence intervals, and
`cosinor.compare()` runs the Bingham / Hotelling \(T^2\) omnibus alongside
circular-aware per-parameter tests. Every metric a batch row or workbook reports is the
same quantity validated in its own article (the [cosinor](cosinor.html) and
[nonparametric](nonparametric.html) families are cross-checked against reference
implementations in the [Validation](validation.html) article), and the meaning of each
column is documented in the [output codebook](output-codebook.html).

## References
