---
title: "Beyond the basics: profile, decomposition, and finer metrics"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
bibliography: ../REFERENCES.bib
csl: ../apa.csl
link-citations: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5, dpi = 110)
```

```{r libraries, message = FALSE, warning = FALSE}
library(actiRhythm)
library(ggplot2)
```

This article continues from the get-started vignette, picking up the same bundled
recording and going past the core workflow. It covers a fuller functional profile of
the day and a singular-spectrum decomposition of the whole series, then a wider set of
methods: finer nonparametric metrics, the shape and phase of the day, time-frequency
and adaptive decomposition, and registration, residual structure, and states. Every
object below is rebuilt from the installed package so the article stands on its own.

```{r data-setup, message = FALSE, warning = FALSE}
agd <- agd.counts(read.agd(example_agd(1), verbose = FALSE))

bin      <- as.integer(as.numeric(agd$timestamp) %/% 600)   # 10-minute bins
counts10 <- as.numeric(tapply(agd$axis1, bin, sum))
t10      <- as.POSIXct(as.numeric(names(tapply(agd$axis1, bin, sum))) * 600,
                       origin = "1970-01-01", tz = "UTC")

state <- sleep.cole.kripke(agd$axis1)
```

## A fuller profile, and a decomposition

The cosinor fits one cosine to the day. `circadian.flm()` fits a functional model of
the same 24-hour profile with several Fourier harmonics, so it follows the real,
asymmetric shape of the day instead of forcing a symmetric curve [@wang2011flm];
the single cosine is its one-harmonic special case.

```{r flm}
flm <- circadian.flm(agd$axis1, agd$timestamp)
flm
```

```{r flm-plot, fig.cap = "The functional fit (line) over the averaged hourly profile (points). The extra Fourier harmonics let the curve follow the asymmetric morning rise and evening decline that a single cosine smooths over."}
ggplot(flm$smooth_curve, aes(t, activity)) +
  geom_point(data = flm$fitted_profile, aes(t, observed), colour = "grey55") +
  geom_line(colour = "#236192", linewidth = 1) +
  labs(x = "Hour of day", y = "Activity") +
  theme_actiRhythm()
```

The extra harmonics buy back the variance the single cosine left on the table, and
the curve now bends through the morning rise the one cosine ran straight past.

`circadian.ssa()` takes the opposite view. Instead of an average day it decomposes
the whole recording with singular spectrum analysis into additive components: a slow
trend, a circadian pair, and noise [@golyandina2013]. The trajectory
matrix grows with the series, so bin a long minute recording to a coarser epoch
first.

```{r ssa}
bin      <- as.integer(as.numeric(agd$timestamp) %/% 600)   # 10-minute bins
counts10 <- tapply(agd$axis1, bin, sum)
t10      <- as.POSIXct(as.numeric(names(counts10)) * 600, origin = "1970-01-01", tz = "UTC")
ssa <- circadian.ssa(as.numeric(counts10), t10)
ssa
```

```{r ssa-plot, fig.height = 3.5, fig.cap = "The circadian component singular spectrum analysis pulls out of the whole recording: the daily rhythm separated from the slow trend and the noise."}
ggplot(data.frame(time = t10, circadian = ssa$circadian), aes(time, circadian)) +
  geom_line(colour = "#236192") +
  labs(x = "Time", y = "Circadian component") +
  theme_actiRhythm()
```

For this recording the circadian pair carries about 14 percent of the variance with a
fundamental period near 24.5 hours, the same slightly-long period the periodogram
found earlier, now isolated as a clean component of the raw series.

## A wider set of methods

The core above answers most questions. actiRhythm also carries a wider set of
circadian methods for finer or more specialized work, shown here on the same
recording.

### Finer nonparametric metrics

`intradaily.variability.multiscale()` averages intradaily variability across bin
sizes, `activity.extrema()` generalizes L5 and M10 to any window with onset and
midpoint times, `dichotomy.index()` measures how cleanly rest separates from the
active day [@mormont2000], and `circadian.daily()` reports each day on its
own so within-recording drift shows. `rest.activity.fragmentation()` adds the
rest and active bout-length view that complements the kRA/kAR rates above.

```{r daily}
daily <- circadian.daily(agd$axis1, agd$timestamp)
dichotomy.index(agd$axis1, rest = state == "S")
```

```{r daily-plot, fig.height = 3, fig.cap = "The M10 onset (the start of the most active ten hours) for each day. Reporting it per day, rather than pooled, shows how steady the active phase is across the recording."}
ggplot(daily$daily, aes(date, M10_onset_h)) +
  geom_line(colour = "grey60") + geom_point(colour = "#236192", size = 2) +
  labs(x = NULL, y = "M10 onset (h)") +
  theme_actiRhythm()
```

### The shape and phase of the day

`cosinor.multicomponent()` adds harmonics to the single cosine and picks how many by
information criterion, `activity.onset.offset()` marks the daily activity onset and
offset by a relative-difference contrast, and
`phase.concentration()` tests whether the daily phase markers cluster (Rayleigh and
Hermans-Rasson).

```{r shape}
cosinor.multicomponent(agd$axis1, agd$timestamp)
phase.concentration(daily$daily$M10_onset_h)
```

### Time-frequency and adaptive decomposition

On the ten-minute series from earlier, `circadian.wavelet()` gives the time-frequency
power spectrum, `ultradian.bandpower()` splits the variance into ultradian bands, and
`circadian.emd()` with `hilbert.huang()` extracts the circadian component
data-adaptively and tracks its instantaneous period.

```{r wavelet-plot, fig.height = 3, fig.cap = "The scale-rectified global wavelet power spectrum (|W|^2/scale, Liu et al. 2007). The peak sits at the circadian period (dashed line at 24 hours), recovered without assuming a fixed cosine shape."}
wav <- circadian.wavelet(as.numeric(counts10), t10, epoch_length = 600)
ggplot(data.frame(period = wav$period_hours, power = wav$global_power),
       aes(period, power)) +
  geom_line(colour = "#236192") +
  geom_vline(xintercept = 24, linetype = 2, colour = "grey50") +
  scale_x_continuous(trans = "log2") +
  labs(x = "Period (h, log scale)", y = "Wavelet power (scale-rectified)") +
  theme_actiRhythm()
```

```{r wavelet-scalogram, fig.height = 3.6, fig.cap = "The full wavelet power scalogram: scale-rectified power across period (log axis) and time, with the cone of influence faded at the ends and the dominant-period ridge traced in white. The circadian band runs the length of the recording."}
plot_wavelet(as.numeric(counts10), t10, epoch_length = 600)
```

```{r emd}
emd <- circadian.emd(as.numeric(counts10), t10, epoch_length = 600)
hilbert.huang(emd)
```

```{r emd-stack, fig.height = 4.5, fig.cap = "The empirical mode decomposition: intrinsic mode functions from finest at the top to the slow trend at the bottom, with the circadian mode (orange) the one nearest 24 hours."}
plot_emd(as.numeric(counts10), t10, epoch_length = 600)
```

### Registration, residual structure, and states

`curve.registration()` aligns the days on their active-phase landmark and reports a
scale-invariant phase marker (the M10-window centre), `residual.spectrum()` examines what the cosinor
leaves behind, and `rest.hmm()` fits a state-space model whose decoded path gives the
probability of rest across the day.

```{r hmm-plot, fig.height = 3, fig.cap = "The 24-hour rest-probability profile from the state-space model: the share of each clock hour the decoded path spends in the rest state."}
hmm <- rest.hmm(as.numeric(counts10), t10)
ggplot(hmm$tod_profile, aes(hour, p_rest)) +
  geom_col(fill = "#236192") +
  labs(x = "Hour of day", y = "P(rest)") +
  theme_actiRhythm()
```

## References
