---
title: "Cosinor models"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
bibliography: ../REFERENCES.bib
csl: ../apa.csl
link-citations: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, dpi = 110)
```

```{r libs, message = FALSE, warning = FALSE}
library(actiRhythm)
library(ggplot2)
```

## The idea, and a rule to remember

The cosinor fits a cosine of fixed period to the rest-activity rhythm and reads
three interpretable numbers off it: the MESOR (the rhythm-adjusted mean), the
amplitude (half the peak-to-trough swing), and the acrophase (the clock time
of the peak). Where the nonparametric metrics make no shape assumption, the cosinor
makes a strong one (a single, symmetric sinusoid) and in return gives you
parameters you can model and compare across people.

The rule to carry through this article: a significant rhythmicity test tells you
the rhythm is not flat, not that a single cosine is the right shape. Always read
the percent-rhythm alongside the p-value; it is how much of the daily structure the
one cosine actually captures, and a low value is the data telling you the shape is
wrong.

## The math

Write the averaged daily activity as \(Y(t)\) at clock time \(t\), with period
\(\tau\) (24 hours by default) and \(\omega = 2\pi/\tau\). The cosinor model is

$$
Y(t) = M + A\cos(\omega t + \phi) + e(t),
$$

which is fitted in its linear form

$$
Y(t) = M + \beta_1\cos(\omega t) + \beta_2\sin(\omega t) + e(t),
$$

so the three parameters come straight from the regression coefficients:

$$
\mathrm{MESOR} = M, \qquad
A = \sqrt{\beta_1^2 + \beta_2^2}, \qquad
\phi = \operatorname{atan2}(-\beta_2,\, \beta_1).
$$

Whether the rhythm is real is the **zero-amplitude F-test** of \(H_0\!: A = 0\)
[@nelson1979], and the share of variance the cosine explains is the
**percent-rhythm**, \(100\,R^2\).

## Assumptions, and when they break

- **One symmetric sinusoid.** The model has a single peak and trough of equal,
  smooth shape. Asymmetric days (a fast morning rise, a slow evening decline),
  squared-off days, naps, and bimodal patterns all violate it. The section
  "When a single cosine is the wrong shape" below makes this concrete.
- **A fixed, known period.** The cosinor assumes \(\tau\); it does not estimate it.
  If the rhythm may be free-running, estimate the period first (see
  [Choosing a method](choosing-a-method.html)).
- **The acrophase is only meaningful once the amplitude is.** A clock time of the
  peak is uninterpretable if the amplitude is not distinguishable from zero.

## Recovering known truth

First, see the fit recover parameters we plant. We build a seven-day recording from
a cosine with a MESOR of 200, an amplitude of 100, and a 16:00 acrophase, then add
noise.

```{r known-truth}
ts <- seq(as.POSIXct("2024-01-01", tz = "UTC"), by = 60, length.out = 7 * 1440)
th <- as.numeric(difftime(ts, ts[1], units = "hours"))
set.seed(1)
known <- pmax(0, 200 + 100 * cos(2 * pi * (th - 16) / 24) + rnorm(length(ts), 0, 15))

ca <- cosinor.analysis(known, ts)
knitr::kable(
  data.frame(parameter = c("MESOR", "amplitude", "acrophase (h)"),
             planted   = c(200, 100, 16),
             recovered = c(ca$mesor, ca$amplitude, ca$acrophase)),
  digits = 2, caption = "The cosinor recovers the planted MESOR, amplitude, and acrophase."
)
```

The rhythmicity test confirms the rhythm is real and reports how much of the
variance the cosine carries.

```{r known-rhythmicity}
rt <- rhythmicity.test(known, ts, cosinor_result = ca)
c(F = rt$F, p_value = rt$p_value, percent_rhythm = rt$percent_rhythm, rhythmic = rt$rhythmic)
```

## On a real recording

The bundled recording runs the same way. `cosinor.analysis()` returns a typed
object; `rhythmicity.test()` checks whether the rhythm is statistically real.

```{r real}
agd <- agd.counts(read.agd(example_agd(1), verbose = FALSE))
cos <- cosinor.analysis(agd$axis1, agd$timestamp, period = 24)
c(mesor = unname(cos$mesor), amplitude = unname(cos$amplitude),
  acrophase = unname(cos$acrophase), r_squared = unname(cos$r_squared))
rhythmicity.test(agd$axis1, agd$timestamp, cosinor_result = cos)
```

```{r fit-plot, fig.cap = "Hourly activity with the fitted 24-hour cosine overlaid. The peak of the curve is the acrophase; its height above the MESOR line is the amplitude. Where the points leave the curve is the structure a single cosine cannot follow."}
plot_extended_cosinor(agd$axis1, agd$timestamp)
```

For the joint uncertainty in amplitude and acrophase together,
`cosinor.confidence.ellipse()` returns the @bingham1982 ellipse; an ellipse that
excludes the origin is the geometric form of "a
detectable rhythm".

```{r ellipse}
el <- cosinor.confidence.ellipse(cos)
c(excludes_origin = el$excludes_origin, rhythm_detected = el$rhythm_detected)
```

## Reading the numbers

- **MESOR** is the rhythm-adjusted mean, the midline of the fitted cosine rather
  than the raw average.
- **Amplitude** is half the peak-to-trough swing; larger is a stronger sinusoidal
  component.
- **Acrophase** is the clock time of the peak and is circular: 23.5 and 0.5 are
  an hour apart, not 23, and it is interpretable only once the amplitude is clearly
  non-zero.
- **Percent-rhythm** is the share of variance the cosine explains; a *low* value
  next to a small p-value is the warning the rule above flags.

## When a single cosine is the wrong shape

This is the rule in action. We build a **siesta day** (an active morning and
evening with a midday dip), which has a clear 24-hour rhythm but is not a single
symmetric cosine.

```{r wrong-way}
hod <- as.numeric(format(ts, "%H")) + as.numeric(format(ts, "%M")) / 60
set.seed(2)
siesta <- pmax(0, 120 + 70 * cos(2 * pi * (hod - 14) / 24) +     # day-night fundamental
                    55 * cos(2 * pi * 2 * (hod - 14) / 24) +     # the midday dip
                    rnorm(length(ts), 0, 12))

cs  <- cosinor.analysis(siesta, ts)
rts <- rhythmicity.test(siesta, ts, cosinor_result = cs)
c(rhythmic = rts$rhythmic, p_value = rts$p_value, percent_rhythm = rts$percent_rhythm)
```

The test is significant (the rhythm is not flat), yet the single cosine
explains only about 62 percent of the day. The missing third is the midday dip,
which a symmetric cosine cannot represent. A multi-harmonic fit recovers it: it
selects two components and explains essentially all of it (the chunk below reports
the R-squared).

```{r wrong-way-fix}
mc <- cosinor.multicomponent(siesta, ts)
c(n_harmonics = mc$n_harmonics, r_squared = mc$r_squared)
mc$harmonics
```

So in practice, let the low percent-rhythm, not the significant p-value, send you to
a richer model rather than a misleading single acrophase.

## The wider cosinor family

- **Asymmetric and squared-off days**: `cosinor.extended()` and
  `cosinor.antilogistic()` fit the sigmoidally transformed cosine of @marler2006,
  adding shape parameters (steepness, up- and down-mesor times)
  the single cosinor lacks.

```{r antilog}
fit <- cosinor.antilogistic(agd$axis1, agd$timestamp)
c(MESOR = fit$MESOR, amplitude = fit$amplitude, acrophase = fit$acrophase,
  UpMesor = fit$UpMesor, DownMesor = fit$DownMesor)
```

- **Several harmonics, chosen objectively**: `cosinor.multicomponent()` adds
  components and picks how many by AIC or BIC, as in the siesta example above.
- **The parametric amplitude as a ratio**: `circadian.quotient()` turns the cosinor
  amplitude and MESOR into the circadian quotient (amplitude / MESOR).
- **Pooling the evidence**: `consensus.rhythmicity()` runs the cosinor F-test, the
  Bingham ellipse, the Lomb-Scargle false-alarm probability, and the chi-square
  periodogram together, and reports a majority vote with a combined p-value.
- **From one subject to a group**: `population.cosinor()` pools per-subject fits
  into a Bingham population-mean rhythm with confidence intervals, and
  `cosinor.compare()` tests whether the rhythm differs between two groups with an
  omnibus Hotelling \(T^2\) and per-parameter tests [@bingham1982]. See
  [Choosing a method](choosing-a-method.html) for when to reach for each.

## Limitations

- **The period is assumed, not estimated.** Set it correctly, or estimate it first
  with the periodograms.
- **A single component is symmetric.** Real days rarely are; read the percent-rhythm
  and escalate to the extended or multicomponent fit when it is low.
- **The acrophase needs a real amplitude.** Report it only when the amplitude is
  clearly distinguishable from zero (the rhythmicity test and the confidence ellipse
  are the checks).

## Reference and validation

The cosinor follows @cornelissen2014 with the zero-amplitude test of @nelson1979,
the anti-logistic extension of @marler2006, and the population framework of
@bingham1982. actiRhythm's single and extended cosinor are cross-checked against
the `cosinor` package and `ActCR::ActExtendCosinor` (to the printed precision)
in the [Validation](validation.html) article and the package's test suite.

## References
