---
title: "Fractal and nonlinear dynamics"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
bibliography: ../REFERENCES.bib
csl: ../apa.csl
link-citations: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, dpi = 110)
```

```{r libs, message = FALSE, warning = FALSE}
library(actiRhythm)
library(ggplot2)
```

## The idea, and a rule to remember

The fractal and nonlinear metrics ask a different question from the cosinor or the
nonparametric battery. Those describe the **shape and depth** of the average day;
the fractal tools describe the **temporal texture** of the minute-to-minute
fluctuations: how this minute's activity remembers the last, and how that memory
stretches across scales from minutes to hours. Healthy activity is neither smooth
nor purely random: it carries a self-similar, "1/f" structure, and that structure
breaks down in aging and disease in ways the amplitude metrics cannot see
[@hu2009].

The rule to carry through this article: **the scaling exponent measures
correlation, not size.** A signal can be large and regular yet have the same
\(\alpha\) as a tiny one; \(\alpha\) reports how the fluctuations are organised
across time, independent of their amplitude. Read it as the rhythm's memory, not
its strength.

## The math

**Detrended fluctuation analysis (DFA).** Start from activity \(x_1,\dots,x_N\) with
mean \(\bar x\), and integrate the mean-centred signal into a profile
[@peng1994; @hu2001]:
$$
Y(k) = \sum_{i=1}^{k} \bigl(x_i - \bar x\bigr).
$$
Split \(Y\) into \(M = \lfloor N/n \rfloor\) non-overlapping windows of length
\(n\), fit and remove a least-squares polynomial (linear by default, quadratic for
DFA-2) inside each window, and pool the residuals into the fluctuation
$$
F(n) = \sqrt{\tfrac{1}{M n}\sum_{k=1}^{M n}\bigl(Y(k) - \widehat{Y}_n(k)\bigr)^2}.
$$
If the signal is self-similar, \(F(n)\) grows as a power law, \(F(n)\propto
n^{\alpha}\), and the **scaling exponent** \(\alpha\) is the slope of \(\log F(n)\)
on \(\log n\). The benchmarks are exact: \(\alpha\approx 0.5\) for uncorrelated
white noise, \(\alpha\approx 1.0\) for \(1/f\) (pink) noise, and \(\alpha\approx 1.5\)
for a Brownian random walk.

**Multifractal DFA (MF-DFA).** A single \(\alpha\) assumes one scaling law governs
the whole signal. MF-DFA relaxes that by weighting the windowed variances
\(F^2(v,n)\) by a moment order \(q\) [@kantelhardt2002]:
$$
F_q(n) = \Bigl[\tfrac{1}{2 N_s}\sum_{v}\bigl(F^2(v,n)\bigr)^{q/2}\Bigr]^{1/q}
        \propto n^{\,h(q)} ,
$$
with windows taken from both ends (hence \(2 N_s\)), the \(q = 0\) case using the
geometric mean, and zero-variance windows dropped.
\(h(2)\) is the ordinary DFA exponent. If \(h(q)\) is flat the signal is
**monofractal**; if \(h(q)\) decreases with \(q\) the signal is **multifractal**, and
the **spectrum width** \(\Delta\alpha\) (the spread of Holder exponents) measures how
much. The width depends on the \(q\) range, so compare it only across recordings
analysed with the same grid (the default is \(q \in [-5, 5]\)).

**Multiscale entropy (MSE).** Complexity is not the same as scaling. Sample entropy
counts how unpredictable a series is: the negative log of the chance that two
sub-sequences matching for \(m\) points still match for \(m+1\) points, within a
tolerance \(r\) [@richman2000]:
$$
\mathrm{SampEn}(m, r) = -\ln \frac{A}{B}.
$$
MSE computes \(\mathrm{SampEn}\) on coarse-grained copies of the series across scales
\(\tau\) [@costa2002]. Noise loses entropy as it is averaged; genuinely complex,
correlated signals hold their entropy across scales.

## Assumptions, and when they break

- **A long, gap-free segment.** Scaling laws need many octaves of window sizes; all
  three functions analyse the **longest continuous non-NA run** and return NA (never
  an error) on series too short to fit a slope. A few days of minute-level data is
  the practical minimum.
- **Stationary fluctuations, not trends.** DFA removes a linear trend inside each
  window, so a slow drift is tolerated, but a strong embedded rhythm or a
  device-off step can masquerade as long-range correlation. Gate on wear time first.
- **Amplitude-blind, by design.** \(\alpha\) is invariant to scaling the signal, so
  it adds to the amplitude metrics rather than replacing them; report it alongside
  the cosinor and nonparametric numbers.
- **Heavy zeros distort the negative moments.** Activity counts have long runs of
  exact zeros, which give near-zero windowed variances. Raised to a negative power
  in MF-DFA these blow up, inflating the spectrum width; the \(q>0\) side and \(h(2)\)
  stay trustworthy.

## Recovering known truth

Before trusting \(\alpha\) on a real recording, watch it recover dynamics whose
answer is fixed by construction. We build three series of the same length and
variance but different memory: **white noise** (no correlation, planted
\(\alpha=0.5\)), **\(1/f\) pink noise** synthesised so its power spectrum falls as
\(1/f\) (planted \(\alpha=1.0\)), and a **Brownian random walk**, the running sum of
white noise (planted \(\alpha=1.5\)).

```{r known-truth}
set.seed(42)
n <- 8192

white <- rnorm(n)                                   # planted alpha = 0.5
brown <- cumsum(rnorm(n))                            # planted alpha = 1.5
pink  <- {                                           # planted alpha = 1.0
  amp  <- 1 / sqrt(1:(n / 2))                        # 1/f amplitude spectrum
  ph   <- runif(n / 2, 0, 2 * pi)
  half <- complex(modulus = c(0, amp), argument = c(0, ph))
  Re(fft(c(half, Conj(rev(half[2:(n / 2)]))), inverse = TRUE))[1:n]
}

planted <- c(white = 0.5, pink = 1.0, brown = 1.5)
recovered <- c(white = fractal.dfa(white)$alpha,
               pink  = fractal.dfa(pink)$alpha,
               brown = fractal.dfa(brown)$alpha)
knitr::kable(
  data.frame(process = names(planted), planted = planted, recovered_alpha = recovered),
  row.names = FALSE, digits = 3,
  caption = "DFA recovers the planted scaling exponent: 0.5 for white noise, 1.0 for 1/f, 1.5 for a random walk."
)
```

The exponent tracks the planted process almost exactly. The fluctuation plot for the
random-walk (brown) series shows the scaling directly: a single straight line on
log-log axes, whose slope is the exponent.

```{r dfa-plot, fig.cap = "DFA fluctuation curve for the random-walk (brown) series on log-log axes. The slope is the scaling exponent (~1.5 here); one straight line is monofractal scaling."}
plot_dfa(brown)
```

## On a real recording

The bundled recording runs the same way. `fractal.dfa()` returns a typed object
that prints its own exponents and an interpretive guide; `alpha1` and `alpha2` split
the short and long timescales at the `breakpoint_min` window (90 minutes by
default).

```{r real}
agd <- agd.counts(read.agd(example_agd(1), verbose = FALSE))
dfa <- fractal.dfa(agd$axis1)
dfa
```

The overall \(\alpha\) sits in the **0.9-1.0 healthy band** (the \(1/f\) signature of
intact activity regulation) rather than near the 0.5 of noise or the 1.5 of a
random walk. That band, and the aging and disease reductions below, were
established by @hu2009 with quadratic detrending (DFA-2); `fractal.dfa()` defaults
to linear DFA-1, so set `detrend_order = 2` to compare directly against those
reference values.

## Reading the numbers

State each exponent in human terms:

- **\(\alpha\)** runs from about 0.5 (uncorrelated) through 1.0 (\(1/f\), the healthy
  target) to 1.5 (random-walk smoothness). Reductions toward 0.5 are reported with
  aging and Alzheimer's disease [@hu2009].
- **\(\alpha_1\) vs \(\alpha_2\)** compare memory below and above the breakpoint; a
  drop from short to long timescales is a crossover, common in activity where the
  circadian band changes the scaling.
- **Spectrum width \(\Delta\alpha\)** (MF-DFA) is near zero for a monofractal and
  grows with multifractality; read it only on the \(q>0\) side for raw counts.
- **Complexity / area** (MSE) is the entropy summed across scales: higher means the
  signal stays unpredictable even when smoothed, the hallmark of physiological
  complexity rather than noise.

A useful pairing is \(\alpha\) with the MSE slope: \(1/f\) scaling and sustained
entropy across scales together say "complex and healthy", which neither number
asserts alone.

## The wider fractal and nonlinear family

**From one exponent to a spectrum.** `mfdfa()` asks whether a *single* scaling law
fits, or whether different-sized fluctuations scale differently. We contrast a
**monofractal** signal (white noise, one law) against a **multifractal** one (a
binomial multiplicative cascade, the textbook multifractal): the spectrum width
separates them cleanly while \(h(2)\) stays interpretable as the DFA exponent.

```{r mfdfa}
set.seed(7)
mono <- rnorm(8192)
cascade <- Reduce(function(v, .) as.numeric(rbind(v * 0.3, v * 0.7)),
                  1:13, accumulate = FALSE)        # binomial multiplicative cascade
multi <- cascade / sd(cascade)

knitr::kable(
  data.frame(
    signal     = c("monofractal (white noise)", "multifractal (cascade)"),
    h2         = c(mfdfa(mono)$alpha_dfa, mfdfa(multi)$alpha_dfa),
    width      = c(mfdfa(mono)$width,     mfdfa(multi)$width)
  ),
  col.names = c("signal", "h(2)", "spectrum width"),
  row.names = FALSE, digits = 3,
  caption = "A narrow spectrum width flags a monofractal; a wide one flags multifractality."
)
```

On the real recording, MF-DFA confirms the DFA exponent through \(h(2)\); the raw
width is inflated by the zero-variance windows noted above, so we report \(h(2)\) as
the reliable summary.

```{r mfdfa-real}
mf <- mfdfa(agd$axis1)
c(h2 = unname(mf$alpha_dfa))
```

```{r mfdfa-plot, fig.height = 3.6, fig.cap = "MF-DFA on the recording: the generalized Hurst exponent h(q) with h(2) dashed, and the singularity spectrum. The wide negative-q side reflects the zero-variance windows noted above, so read h(2) and the q > 0 side."}
plot_mfdfa(agd$axis1)
```

**Complexity across scales.** `multiscale.entropy()` applies sample entropy to
coarse-grained copies of the recording, separating signals that stay complex when
smoothed from noise that loses entropy with scale.

```{r mse}
mse <- multiscale.entropy(agd$axis1, scales = 1:20)
mse
```

```{r mse-plot, fig.height = 3.4, fig.cap = "Sample entropy across coarse-graining scales. A flat or rising curve is the signature of physiological complexity; a curve that falls toward zero is noise-like."}
plot_mse(agd$axis1, scales = 1:20)
```

The complexity index (`area`, the entropy summed across scales) condenses that curve
to a single number for comparison between recordings.

## Limitations

- **Data hunger.** Scaling exponents are unstable on short or gappy series; the
  functions fall back to the longest clean run and return NA when even that is too
  short. Treat \(\alpha\) from under a few days with caution.
- **Embedded rhythm leaks into the slope.** A strong 24-hour component adds
  curvature to the log-log plot near the daily scale; the \(\alpha_1/\alpha_2\) split
  exposes it, but a single pooled \(\alpha\) can hide it.
- **Negative-moment fragility.** MF-DFA spectrum width on raw activity counts is
  inflated by zero-variance windows; lean on \(h(2)\) and the \(q>0\) side, or
  pre-process the zeros, before reporting \(\Delta\alpha\).
- **Complementary, not diagnostic.** A reduced \(\alpha\) is a marker, not a
  diagnosis; interpret it next to the amplitude, timing, and fragmentation metrics,
  never in isolation.

## Reference and validation

DFA follows @peng1994, with trend and crossover handling from @hu2001 and the
activity-specific framing of @hu2009; the
multifractal generalisation is @kantelhardt2002; and multiscale entropy combines the
sample-entropy estimator of @richman2000 with the multiscale construction of
@costa2002. actiRhythm's DFA and
MSE are implemented in base R against these reference algorithms and exercised
against their known limits (\(\alpha\to0.5\) for white noise, \(\to1.5\) for a
random walk, entropy declining with scale for noise) in the package's test suite;
see also the [Validation](validation.html) article.

## References
