---
title: "Reading data and wear-time detection"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
bibliography: ../REFERENCES.bib
csl: ../apa.csl
link-citations: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, dpi = 110)
```

```{r libs, message = FALSE, warning = FALSE}
library(actiRhythm)
library(ggplot2)
```

## The idea, and a rule to remember

Every rhythm metric in actiRhythm starts from the same two columns: a regular
**timestamp** and a per-epoch **count**. Getting your recording into that shape,
and marking the stretches where the device was not on the body, is the whole job
of this article. Whether the data arrives as an ActiGraph `.agd` database, an
ActiLife `.csv` export, or a column in some other data frame, the reader returns
the same tidy frame, and the analysis functions never need to know where it came
from.

The rule to carry through this article: **a count of zero is not the same as rest.**
A taken-off device reads as a long, flat block of zeros that looks like the deepest
sleep you will ever measure, and if you feed it to the rhythm functions as if it
were real rest, it will distort every number. Detecting non-wear and passing the
result as `wear_time` is how you keep "the device was in a drawer" from
masquerading as "the subject was perfectly still".

## The shape every reader returns

There is little math here; the contract is structural. Each entry point hands back a
data frame whose first column is a POSIXct `timestamp` on a fixed epoch grid and
whose remaining columns are counts:

- `read.agd()` opens the SQLite `.agd` database and returns its raw tables;
  `agd.counts()` turns that into the tidy `timestamp` / `axis1` / `axis2` /
  `axis3` / `steps` frame (plus `lux` and inclinometer columns when present).
- `read.actigraph.csv()` parses an ActiLife epoch CSV (start time and epoch
  length come from the header, the count columns are matched by name) and returns
  the **same shape** directly.
- `counts.from.data.frame()` pulls a count column (and optional time column) out of
  any data frame, for non-ActiGraph series.

Because the shape is shared, a series read three different ways flows into
`circadian.rhythm()`, `cosinor.analysis()`, and the rest without a single change.
Reading **raw** acceleration (`.gt3x`, `.cwa`, GENEActiv `.bin`) and computing
counts from it is a larger topic with its own filters and calibration; that lives in
the [raw pipeline](raw-pipeline.html) article. Here we stay at the epoch level.

## Reading, and when it breaks

- **A real timestamp is mandatory.** `agd.counts()` fails fast if neither a
  `dataTimestamp` nor a `timestamp` column is present, because a synthesized
  sequential index silently breaks every day-level and circadian computation. Fix
  the source rather than inventing time.
- **The epoch grid must be regular.** IS, IV, and the non-wear runs are all defined
  on evenly spaced epochs. `read.actigraph.csv()` builds the grid from the header's
  epoch length; if that header line is missing it falls back to `epoch_length` (60 s
  by default), so check it when the CSV is unusual.
- **Date formats vary.** ActiLife writes `M/d/yyyy` by default, but exports made on
  a non-US locale may differ; pass `date_format` to `read.actigraph.csv()` to match.
- **Zero is ambiguous.** A zero count can be genuine stillness or a device that is
  off. Nothing in the reading step can tell them apart. That is what the non-wear
  algorithms below are for.

## Recovering known truth: an injected device-off block

Before trusting non-wear detection on real data, plant a gap whose location we know
and confirm the algorithms recover *exactly* that block. We build a clean three-day
count series from a day-night cosine, then switch the device "off" (a run of
zeros) for a known 180-minute window.

```{r plant}
set.seed(1)
n   <- 3 * 1440
ts  <- seq(as.POSIXct("2024-01-01", tz = "UTC"), by = 60, length.out = n)
hod <- as.numeric(format(ts, "%H")) + as.numeric(format(ts, "%M")) / 60
worn <- pmax(0, round(100 + 80 * cos(2 * pi * (hod - 14) / 24) + rnorm(n, 0, 15)))

gap <- 1000:1179            # a 180-minute device-off block at a known location
counts <- worn; counts[gap] <- 0
```

Both detectors return a logical mask (`TRUE` = wear, `FALSE` = non-wear), one
value per epoch. We ask each which epochs it flagged as non-wear and compare to the
planted block.

```{r recover}
choi <- detect.nonwear.choi(counts)
tro  <- detect.nonwear.troiano(counts)

recovered <- function(mask) { idx <- which(!mask); c(first = min(idx), last = max(idx), n = length(idx)) }
knitr::kable(
  rbind(planted  = c(first = min(gap), last = max(gap), n = length(gap)),
        choi     = recovered(choi),
        troiano  = recovered(tro)),
  caption = "Both algorithms recover the planted device-off block exactly: epochs 1000-1179."
)
```

Each algorithm flags the same 180 epochs we silenced, to the epoch. The figure makes
it visual: the shaded band is the recovered non-wear mask, sitting exactly over the
flat-line gap.

```{r mask-plot, fig.height = 3, fig.cap = "A window of the planted series. The flat zero block is the injected device-off period; the shaded band is the Choi non-wear mask, recovered exactly over it."}
w  <- 900:1280
df <- data.frame(epoch = w, counts = counts[w], nonwear = !choi[w])
ggplot(df, aes(epoch, counts)) +
  geom_rect(data = df[df$nonwear, ],
            aes(xmin = epoch - 0.5, xmax = epoch + 0.5, ymin = -Inf, ymax = Inf),
            fill = "#E8A33D", alpha = 0.4, inherit.aes = FALSE) +
  geom_line(colour = "#236192") +
  labs(x = "epoch (minute)", y = "counts") +
  theme_actiRhythm()
```

**The two algorithms are not identical, though.** They agree on this clean block but
diverge by design, and the differences matter on real data. Choi requires a long
window and guards a tolerated spike with flanking all-zero windows; Troiano uses a
shorter default window and a count *ceiling* on spikes [@choi2011; @troiano2008].
Two short experiments expose the gap.

```{r contrast}
# (1) a borderline 75-minute gap: longer than Troiano's 60-min frame, shorter than Choi's 90
short <- worn; short[1000:1074] <- 0
c75 <- c(choi = sum(!detect.nonwear.choi(short)), troiano = sum(!detect.nonwear.troiano(short)))

# (2) the same 180-min gap, broken by a single high-count spike of 250
spike <- worn; spike[gap] <- 0; spike[1090] <- 250
csp <- c(choi = sum(!detect.nonwear.choi(spike)), troiano = sum(!detect.nonwear.troiano(spike)))

knitr::kable(rbind(`75-min gap` = c75, `gap with 250-count spike` = csp),
             caption = "Non-wear epochs flagged. The two algorithms disagree where their defaults differ.")
```

The 75-minute gap clears Troiano's 60-minute frame but not Choi's 90-minute one, so
Troiano flags it and Choi does not. The lone 250-count spike sits below nothing Choi
cares about (it is flanked by zeros, so Choi absorbs it and keeps the full
180-epoch block), while it exceeds Troiano's `stoplevel` of 100, ending the bout at
that minute and leaving 179. Neither is "right"; they encode different definitions of
non-wear, and you choose to match the literature you are comparing against.

The recovered mask is exactly what the analysis functions want. Pass it as
`wear_time` and the rhythm metrics are computed only over worn epochs:

```{r feeds}
r <- circadian.rhythm(counts, ts, wear_time = choi)
c(IS = round(r$IS, 3), valid = isTRUE(!is.null(r$IS)))
```

## On a real recording

The bundled recording runs the same path: open the `.agd`, extract counts, detect
non-wear, gate the analysis on the result.

```{r real}
agd <- agd.counts(read.agd(example_agd(1), verbose = FALSE))
c(epochs = nrow(agd), columns = ncol(agd))
names(agd)
```

```{r real-nonwear}
wt <- detect.nonwear.choi(agd$axis1)
c(epochs = length(wt), worn = sum(wt), nonwear = sum(!wt),
  pct_worn = round(100 * mean(wt), 1))
```

This file carries inclinometer columns, which some sleep routines can use; a quick
check tells you whether they are present before you reach for them.

```{r inclin}
has.inclinometer(read.agd(example_agd(1), verbose = FALSE))
```

Feeding the wear mask forward is the same one argument as before. The metrics now
ignore any device-off stretch.

```{r real-feed}
cr <- circadian.rhythm(agd$axis1, agd$timestamp, wear_time = wt)
c(IS = round(cr$IS, 3), IV = round(cr$IV, 3), RA = round(cr$RA, 3))
```

## Reading the numbers

Wear-time output is plain, but the yardsticks matter:

- **Percent worn.** A logical mask; its mean is the fraction of the recording on the
  body. Many analyses ask for a minimum number of valid hours per day before a day
  counts. `circadian.rhythm()` enforces that through `min_valid_hours`.
- **Non-wear runs, not scattered epochs.** Both algorithms flag *runs* of at least
  `frame` minutes, so the mask is blocky by construction; isolated zeros inside an
  active day stay "wear". That is intentional: a single quiet minute is rest, not a
  removed device.
- **Choi versus Troiano.** Choi's longer frame and flanking-window guard make it
  conservative (it flags less, more confidently); Troiano's shorter frame and count
  ceiling make it quicker to call non-wear. Report which you used, and its
  parameters.

## The wider input/output family

The same tidy frame is the hub for several entry and exit points.

**An ActiLife CSV, no database needed.** `read.actigraph.csv()` reads an epoch CSV
export and returns the `timestamp`/count frame directly, computing the vector
magnitude when the three axes are present.

```{r csv}
tf  <- tempfile(fileext = ".csv")
hdr <- c(
  "------------ Data File Created By ActiGraph wGT3X-BT ActiLife v6.13.4 date format M/d/yyyy at 60 sec ------------",
  "Serial Number: MOS2A12345678",
  "Start Time 00:00:00",
  "Start Date 1/1/2024",
  "Epoch Period (hh:mm:ss) 00:01:00",
  "--------------------------------------------------",
  "Axis1,Axis2,Axis3,Steps")
set.seed(2)
body <- sprintf("%d,%d,%d,%d", sample(0:500, 6), sample(0:300, 6),
                sample(0:300, 6), sample(0:20, 6))
writeLines(c(hdr, body), tf)

csv <- read.actigraph.csv(tf)
names(csv)
head(csv, 3)
```

**Any data frame at all.** `counts.from.data.frame()` is the device-neutral door:
name the count and time columns and it returns `timestamp`/`counts`, ready for the
analysis functions.

```{r dataframe}
df <- data.frame(activity = c(0, 50, 300),
                 clock    = c("2024-01-01 00:00:00", "2024-01-01 00:01:00",
                              "2024-01-01 00:02:00"))
counts.from.data.frame(df, count_col = "activity", time_col = "clock")
```

**Writing back out.** `write.agd()` saves a counts frame as a standard `.agd`
SQLite database (the same schema `read.agd()` opens), so you can round-trip,
share, or hand a cleaned series to ActiLife.

```{r roundtrip}
out  <- tempfile(fileext = ".agd")
write.agd(agd, out)
back <- agd.counts(read.agd(out, verbose = FALSE))
c(rows_in = nrow(agd), rows_out = nrow(back),
  axis1_identical = isTRUE(all.equal(round(agd$axis1), back$axis1)))
```

**From a raw `.gt3x`, in one step.** When you have a raw file rather than epoch
counts, `gt3x.to.agd()` computes ActiGraph counts (the agcounts implementation of
@neishabouri2022) and writes them straight to an `.agd`,
carrying the device and subject metadata from the header. It needs the optional
`agcounts` and `read.gt3x` packages, so it is shown but not run here.

```{r gt3x, eval = FALSE}
agd_path <- gt3x.to.agd("subject01.gt3x", epoch = 60)
agd      <- agd.counts(read.agd(agd_path))
```

## Limitations

- **Reading cannot disambiguate zeros.** The reader trusts the file; only the
  non-wear step separates a stationary device from genuine rest, and only as well as
  its parameters are chosen.
- **Defaults are conventions, not truths.** Choi's 90-minute and Troiano's 60-minute
  frames come from validation studies on hip-worn devices; wrist data and other
  populations may warrant different `frame`, `spike_tolerance`, and `stream` /
  `stoplevel` values.
- **CSV header parsing is best-effort.** A heavily customized ActiLife export may
  hide the epoch length or use an unexpected date format; verify the first timestamp
  and epoch spacing after reading, and set `date_format`/`epoch_length` if needed.
- **`.agd` writing is the count schema, not the full ActiLife object.**
  `write.agd()` stores `settings` and `data`; it does not reconstruct sleep,
  wear-time-validation, or capsense tables an original device file might carry.

## Reference and validation

The two non-wear algorithms implement @choi2011 and the
@troiano2008 NHANES protocol; `gt3x.to.agd()` and the raw
readers compute counts with the open ActiGraph-count algorithm of
@neishabouri2022, the same family validated against ActiLife by
@brond2017. actiRhythm's non-wear masks and count round-trips are
cross-checked against reference implementations in the [Validation](validation.html)
article and the package's test suite; the raw count computation is documented in the
[raw pipeline](raw-pipeline.html) article.

## References
