---
title: "Nonparametric rhythm metrics"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
bibliography: ../REFERENCES.bib
csl: ../apa.csl
link-citations: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, dpi = 110)
```

```{r libs, message = FALSE, warning = FALSE}
library(actiRhythm)
library(ggplot2)
```

## The idea, and a rule to remember

The nonparametric metrics describe a rest-activity rhythm **without assuming it
has any particular shape**. Where the cosinor asks "how well does a cosine fit?",
interdaily stability, intradaily variability, and relative amplitude ask only "how
regular, how fragmented, and how deep is the rhythm?", questions you can answer
even when the daily profile is asymmetric, bimodal, or squared-off, which real
rest-activity data usually is.

The rule to carry through this article: **nonparametric metrics make no shape
assumption, but they are sensitive to epoch length and to missing data.** Two
recordings binned to different epochs are not directly comparable, and a few
device-off days can move the numbers as much as a real change in the rhythm.

## The math

Bin the recording to hourly means: let \(x_1, \dots, x_n\) be the \(n\) hourly
values with overall mean \(\bar x\), and let \(\bar x_h\) be the mean across all
days at hour-of-day \(h\), over \(p = 24\) hours. The IS and IV formulas below are
computed on this hourly series; the raw-epoch series is used separately for the
multiscale and fractal measures.

**Interdaily stability (IS)**, how tightly the pattern repeats from one day to
the next [@witting1990]:
$$
\mathrm{IS} = \frac{n \sum_{h=1}^{p}(\bar x_h - \bar x)^2}{p \sum_{i=1}^{n}(x_i - \bar x)^2}.
$$

**Intradaily variability (IV)**, how fragmented the rhythm is, from the squared
hour-to-hour differences [@witting1990]:
$$
\mathrm{IV} = \frac{n \sum_{i=2}^{n}(x_i - x_{i-1})^2}{(n-1)\sum_{i=1}^{n}(x_i - \bar x)^2}.
$$

**Relative amplitude (RA)**, the day-night contrast, from the most-active 10-hour
window M10 and the least-active 5-hour window L5 of the average day
[@vansomeren1999]:
$$
\mathrm{RA} = \frac{\mathrm{M10} - \mathrm{L5}}{\mathrm{M10} + \mathrm{L5}}, \qquad
\mathrm{M10} = \max_s \overline{x}_{[s,\,s+10\text{h}]}, \quad
\mathrm{L5} = \min_s \overline{x}_{[s,\,s+5\text{h}]} .
$$

`circadian.rhythm()` returns all of these, plus the 1-hour extremes L1 and M1, the
autocorrelation predictability \(\phi\), and the onset times of each window.

The L5 and M10 windows are placed on a minute-resolution average day and reported in
counts per minute, the sliding-window convention also used by nparACT and GGIR. These
values track but are not identical to the original hourly counts-per-hour of
@witting1990, and they depend on epoch length.

## Assumptions, and when they break

- **Evenly spaced epochs.** IS and IV are defined on a regular grid; gaps must be
  handled (the package works on the valid epochs and reports coverage).
- **Enough whole days.** IS is a between-day quantity, so a one- or two-day record
  gives an unstable estimate. `circadian.rhythm()` sets aside days with too little
  valid recording before computing.
- **A fixed epoch length.** IV in particular rises as the epoch shortens (finer
  sampling exposes more hour-to-hour change), so only compare IV computed at the
  same epoch. The multiscale tools below make that dependence explicit.
- **Wear time.** A stationary, taken-off device reads as deep rest; gate the counts
  on valid wear first if the recording has device-off stretches.

## Recovering known truth

Before trusting these numbers on real data, it is worth seeing them behave on data
whose answer we know. We build three seven-day recordings: one whose day repeats
**exactly**, one whose peak **wanders** a few hours from day to day, and one that
is **pure noise**. IS should fall from near 1 toward 0 across the three, and IV
should rise toward its noise ceiling of about 2.

```{r known-truth}
ts  <- seq(as.POSIXct("2024-01-01", tz = "UTC"), by = 60, length.out = 7 * 1440)
hod <- as.numeric(format(ts, "%H")) + as.numeric(format(ts, "%M")) / 60
day <- (seq_along(ts) - 1) %/% 1440
set.seed(1)

regular  <- pmax(0, 100 + 80 * cos(2 * pi * (hod - 14) / 24)) + rnorm(length(ts), 0, 5)
acro_jit <- (14 + rnorm(7, 0, 4))[day + 1]                      # acrophase wanders by day
jittered <- pmax(0, 100 + 80 * cos(2 * pi * (hod - acro_jit) / 24)) + rnorm(length(ts), 0, 5)
noise    <- pmax(0, rnorm(length(ts), 100, 80))

recover <- function(x) {
  r <- circadian.rhythm(x, ts)
  c(IS = r$IS, IV = r$IV, RA = r$RA)
}
knitr::kable(
  rbind(regular = recover(regular), jittered = recover(jittered), noise = recover(noise)),
  digits = 3,
  caption = "IS falls and IV rises as a planted rhythm is made irregular, then noise."
)
```

A rhythm that repeats exactly returns IS at its ceiling of 1 with a near-zero IV;
day-to-day phase jitter pulls IS down to a third of that while RA shrinks; pure
noise leaves IS near 0 and IV at its ceiling near 2. The metrics report exactly the
regularity we put in.

## On a real recording

The recording bundled with the package runs the same way. `circadian.rhythm()`
returns a typed object that prints its own metrics.

```{r real}
agd <- agd.counts(read.agd(example_agd(1), verbose = FALSE))
cr  <- circadian.rhythm(agd$axis1, agd$timestamp)
cr
```

```{r actogram, fig.height = 5, fig.cap = "Double-plotted actogram of the bundled recording. The active rows do not stack into one vertical band; they drift, which is the low interdaily stability the metrics report."}
plot_actogram(agd$axis1, agd$timestamp)
```

The average day makes the L5 and M10 windows concrete: the least-active 5-hour block
and the most-active 10-hour block of the folded profile.

```{r profile, fig.height = 3.5, fig.cap = "The averaged 24-hour profile (mean with a one-SD band), shading the least-active 5-hour window L5 and the most-active 10-hour window M10."}
plot_profile(agd$axis1, agd$timestamp)
```

## Reading the numbers

State each metric in human terms:

- **IS** runs 0 to 1; near 1 is a highly stable 24-hour pattern, below about 0.3 is
  weak. Here it is low: the days do not land at the same clock time.
- **IV** runs from about 0 (a smooth sine) to about 2 (noise); above 2 flags a strong
  ultradian component or a too-short record. The denominator is the data's own
  variance, so IV is amplitude-independent.
- **RA** runs 0 to 1; higher is a stronger day-night contrast.

The most informative reading is often the **combination**. This recording pairs a
high RA (near 0.98) with a low IS (near 0.23): the days are strongly active, but the
pattern does not repeat at the same clock time, a strong rhythm carried on
irregular timing, which neither number says on its own.

## The wider nonparametric family

The same averaged-profile idea generates a family of related descriptors.

**Generalising L5/M10 to any window.** `activity.extrema()` reports the least- and
most-active window of any length, with onset and midpoint times. L1/M1 and L5/M10
are special cases.

```{r extrema}
activity.extrema(agd$axis1, agd$timestamp, windows = c(1, 5, 10))$table
```

**Making the epoch dependence explicit.** Because IV and IS depend on the bin size,
`intradaily.variability.multiscale()` and `circadian.is.multiscale()` recompute them
across a range of epochs and average them into a single IVm and ISm, which vary less
across recordings than the hourly values [@goncalves2014]. IVm spans 1 to 60-minute
bins; ISm uses the bin sizes that divide 1440, so each day splits into whole bins.

```{r multiscale}
intradaily.variability.multiscale(agd$axis1, agd$timestamp)
circadian.is.multiscale(agd$axis1, agd$timestamp)
```

```{r multiscale-plot, fig.height = 3, fig.cap = "IS and IV across epoch lengths, with the averaged ISm and IVm (dashed). Both vary less across recordings than a single hourly value."}
plot_multiscale(agd$axis1, agd$timestamp)
```

**How cleanly rest separates from activity.** The dichotomy index I<O is the
percentage of rest-span counts that fall below the median of the active-span counts
[@mormont2000]:
$$
I{<}O = 100 \times \frac{\#\{\, x_i < \tilde{x}_{\text{active}} : i \in \text{rest} \,\}}{N_{\text{rest}}},
$$
where \(\tilde{x}_{\text{active}}\) is the median active-span count. A high I<O means
rest is quiet relative to the active day. Here we mark the night hours as rest.

```{r dichotomy}
h <- as.numeric(format(agd$timestamp, "%H"))
dichotomy.index(agd$axis1, rest = h >= 23 | h < 7)
```

```{r dichotomy-plot, fig.height = 3.5, fig.cap = "Rest-span and active-span count distributions. A high I<O means most rest-span counts fall below the active-span median (dashed)."}
plot_dichotomy(agd$axis1, rest = h >= 23 | h < 7)
```

**Fragmentation.** `state.transitions()` gives the kRA and kAR rates, the
rest-to-active and active-to-rest probabilities read off the plateau of the
bout-length survival curve [@lim2011]. `transition.probability()` gives the
maximum-likelihood and Bayesian transition probabilities of @danilevicz2024, taken
straight from the bout counts. Both capture fragmentation a single amplitude cannot.

```{r fragmentation}
state.transitions(agd$axis1)
transition.probability(agd$axis1)[c("tp_ra_mle", "tp_ar_mle")]
```

```{r fragmentation-plot, fig.height = 3.5, fig.cap = "Transition hazards versus bout length, with a LOWESS fit and the sustained rates kRA and kAR (dashed)."}
plot_transitions(agd$axis1)
```

**Per day, to see drift.** `circadian.daily()` reports each day on its own, so
within-recording change shows instead of being averaged away.

```{r daily, fig.height = 3, fig.cap = "The most-active-window onset for each day. Reporting it per day, rather than pooled, shows how steady the active phase is."}
daily <- circadian.daily(agd$axis1, agd$timestamp)$daily
ggplot(daily, aes(date, M10_onset_h)) +
  geom_line(colour = "grey60") + geom_point(colour = "#236192", size = 2) +
  labs(x = NULL, y = "M10 onset (h)") +
  theme_actiRhythm()
```

## Limitations

- **Missing data moves the numbers.** A few device-off days can shift IS as much as
  a real change; always check coverage and gate on wear time.
- **Epoch dependence.** Compare IV (and to a lesser extent IS) only across records
  binned to the same epoch; use the multiscale view when epochs differ.
- **A minimum of valid days.** Very short records give unstable IS; the function
  enforces a per-day validity threshold and reports how many days qualified.

## Reference and validation

The nonparametric battery follows @witting1990 and @vansomeren1999, with the
multiscale and dichotomy extensions of @goncalves2014 and @mormont2000, and the
fragmentation rates of @lim2011 and @danilevicz2024.
actiRhythm's IS, IV, RA, L5, and M10 are cross-checked against the
`nparACT` and `ActCR` reference implementations (to the printed precision) in
the [Validation](validation.html) article and the package's test suite.

## References
