---
title: "Nonstationary and complex rhythms"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
bibliography: ../REFERENCES.bib
csl: ../apa.csl
link-citations: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, dpi = 110)
```

```{r libs, message = FALSE, warning = FALSE}
library(actiRhythm)
library(ggplot2)
```

The getting-started vignette runs every method on one steady recording, where a single
cosine and one set of nonparametric numbers describe the rhythm well. Real recordings
are rarely that tidy. A rhythm can drift in period, advance or delay its phase partway
through, or fragment as the rest period breaks up. When any of that happens, a summary
that collapses the whole recording to one number averages the change away. This vignette
builds a recording that is not stationary and works through the methods made to see what
a static summary misses.

## A recording that shifts

There is no nonstationary file to bundle, so we synthesise one: fourteen days at
one-minute epochs, with an active block by day and quiet nights, whose peak advances
about two and a half hours earlier across day 7, and whose nights fragment more as the
recording goes on. A few methods that locate per-day landmarks read the minute series
directly; the heavier decompositions read a ten-minute binned copy, the same split the
getting-started vignette uses.

```{r make-data}
set.seed(42)
days <- 14
t    <- as.POSIXct("2024-03-04", tz = "UTC") + (seq_len(days * 1440) - 1) * 60
hour <- as.numeric(format(t, "%H")) + as.numeric(format(t, "%M")) / 60
day  <- (seq_along(t) - 1) %/% 1440 + 1

acro  <- 15 - 2.5 * pmin(pmax(day - 7, 0), 1)            # peak advances over day 7
block <- plogis(4 * cos(2 * pi * (hour - acro) / 24))    # active by day, quiet at night
frag  <- (day - 1) / (days - 1)                          # 0 to 1 across the recording
counts <- rpois(length(t), 5 + 300 * block)
night_break <- runif(length(t)) < 0.10 * frag & block < 0.2
counts[night_break] <- rpois(sum(night_break), 150)      # nights fragment late on

agg      <- tapply(counts, as.integer(as.numeric(t) %/% 600), sum)   # 10-minute bins
counts10 <- as.numeric(agg)
t10      <- as.POSIXct(as.numeric(names(agg)) * 600, origin = "1970-01-01", tz = "UTC")
```

The double-plotted actogram tells the story at a glance. The active band sits at the
same clock time for the first week, then steps to the left around day 7 as the peak
moves earlier, and the later rows scatter as the nights fragment.

```{r actogram, fig.height = 5, fig.cap = "Double-plotted actogram of the synthetic recording. The activity band advances about two and a half hours earlier around day 7, and the later days fragment. A single cosine fit to all fourteen days cannot represent either change."}
plot_actogram(counts, t, scale = "sqrt")
```

## What a static summary misses

The usual descriptors still run, and they still return one answer for the whole
recording. The cosinor averages all fourteen days onto a single 24-hour profile and
fits one cosine to it, so the two phase regimes blur into one smeared peak. The
nonparametric metrics give one interdaily-stability and one intradaily-variability
number: fragmentation pushes them toward a less stable, more variable rhythm, but that
single value is a compromise between the clean early days and the broken late ones, and
neither number says that the phase moved.

```{r static}
cosinor.analysis(counts, t, period = 24)
rhythm <- circadian.rhythm(counts, t)
c(IS = rhythm$IS, IV = rhythm$IV, RA = rhythm$RA)
```

## Seeing the drift over time

`circadian.spectrogram()` slides a window across the recording and computes a chi-square
periodogram [@sokolove1978] in each, so the dominant period and its strength become a surface over time
rather than a single estimate. The circadian band stays near 24 hours here, but its
power tracks the regime change instead of averaging through it.

```{r spectrogram, fig.height = 4, fig.cap = "Sliding-window periodogram. Each column is one 48-hour window stepped every 12 hours; colour is periodogram power. The circadian band is visible throughout, and its strength shifts as the recording moves from the stable first week into the fragmented second."}
circadian.spectrogram(counts10, t10, window_hours = 48, step_hours = 12,
                      epoch_length = 600)$plot
```

## Locating the shift

When the question is *when* the pattern changed rather than how, `sleep.changepoints()`
finds the per-night rest onsets and wake times directly from the activity, anchored to a
cosinor threshold [@chensun2024]. Because the active peak advances, the detected sleep onsets move
earlier in the second half of the recording.

```{r changepoints, fig.height = 3.2, fig.cap = "Detected sleep onset for each night, as a clock hour. The onset holds steady through the first week and then jumps earlier once the active phase advances, which the spectrogram and the actogram show from different angles."}
cp <- sleep.changepoints(counts, t)
ep <- cp$sleep_episodes
ep$onset_hour <- as.numeric(format(ep$sleep_onset, "%H")) +
                 as.numeric(format(ep$sleep_onset, "%M")) / 60
ggplot(ep, aes(seq_along(onset_hour), onset_hour)) +
  geom_line(colour = "grey60") + geom_point(colour = "#236192", size = 2) +
  labs(x = "Night", y = "Sleep onset (clock hour)") +
  theme_actiRhythm()
```

## Decomposing the signal

Singular spectrum analysis separates the series into additive pieces without assuming a
shape: a slow trend, the circadian oscillation, and a residual [@golyandina2013; @vautard1992]. `circadian.ssa()` returns
each piece on the original time axis, so the circadian component can be read on its own
while the trend and residual carry the parts the cosinor would have forced into one curve.

```{r ssa, fig.height = 5, fig.cap = "The singular-spectrum decomposition. The circadian component holds the 24-hour rhythm and bends with the phase advance, the trend follows the slow drift in overall activity, and the residual collects the rising fragmentation."}
ssa <- circadian.ssa(counts10, t10)
parts <- rbind(
  data.frame(t = t10, value = ssa$circadian, part = "circadian"),
  data.frame(t = t10, value = ssa$trend,     part = "trend"),
  data.frame(t = t10, value = ssa$residual,  part = "residual"))
parts$part <- factor(parts$part, c("circadian", "trend", "residual"))
ggplot(parts, aes(t, value)) +
  geom_line(colour = "#236192", linewidth = 0.3) +
  facet_wrap(~part, ncol = 1, scales = "free_y") +
  labs(x = NULL, y = NULL) +
  theme_actiRhythm()
```

Empirical mode decomposition reaches the same idea from the other direction [@huang1998]. It pulls out
data-adaptive oscillatory modes, and `hilbert.huang()` turns the circadian mode into an
instantaneous period, one value per epoch, so a period that wanders shows up as a moving
line rather than a single number.

```{r emd, fig.height = 3.2, fig.cap = "Instantaneous period of the circadian mode from the Hilbert-Huang transform. It stays near 24 hours but is not flat: the phase advance leaves its mark on the cycle-by-cycle estimate."}
emd <- circadian.emd(counts, t)
hht <- hilbert.huang(emd)
ggplot(data.frame(t = emd$times, period = hht$period), aes(t, period)) +
  geom_line(colour = "#236192") +
  geom_hline(yintercept = 24, linetype = 2, colour = "grey50") +
  coord_cartesian(ylim = c(18, 30)) +
  labs(x = NULL, y = "Instantaneous period (h)") +
  theme_actiRhythm()
```

## Quantifying the complexity

Fragmentation is a change in structure, not just in level, so it shows up in the
multiscale measures. Splitting the recording into its clean first half and its broken
second half, the late half carries more sample entropy [@costa2002] and a narrower multifractal
spectrum [@kantelhardt2002], both signs that its dynamics have simplified toward noise: less predictable
from epoch to epoch, and less richly self-similar across scales.

```{r complexity}
half  <- length(counts10) %/% 2
early <- counts10[1:half]; late <- counts10[(half + 1):length(counts10)]
mse_e <- multiscale.entropy(early, scales = 1:10)
mse_l <- multiscale.entropy(late,  scales = 1:10)
c(early_complexity = mse_e$area, late_complexity = mse_l$area)
c(early_mf_width = mfdfa(early)$width, late_mf_width = mfdfa(late)$width)
```

```{r mse-plot, fig.height = 3.2, fig.cap = "Sample entropy against coarse-graining scale for the two halves. The fragmented second half carries more entropy, a less predictable and more broken rhythm."}
mse <- rbind(data.frame(scale = mse_e$scales, sampen = mse_e$mse, half = "first half"),
             data.frame(scale = mse_l$scales, sampen = mse_l$mse, half = "second half"))
ggplot(mse, aes(scale, sampen, colour = half)) +
  geom_line() + geom_point(size = 1.5) +
  scale_colour_manual(values = c("first half" = "#236192", "second half" = "#E69F00")) +
  labs(x = "Scale factor", y = "Sample entropy", colour = NULL) +
  theme_actiRhythm()
```

## A phase that survives the drift

When the phase moves, the averaged acrophase is a poor summary of any single day.
`curve.registration()` instead finds each day's own activity landmark, aligns the days on
it, and reports a chronotype phase plus how much that landmark varies [@krafty2019]. The variability is
large here precisely because the phase advanced, which is the honest answer.

```{r registration}
reg <- curve.registration(counts, t)
c(mean_M10_hour = reg$mean_M10, phase_variability_h = reg$phase_sd)
```

```{r reg-plot, fig.height = 3.2, fig.cap = "The most-active-window centre for each day. It is steady through the first week and then jumps earlier, so the across-day phase variability that registration reports is real structure, not noise."}
ggplot(reg$landmarks, aes(seq_along(M10_center_h), M10_center_h)) +
  geom_line(colour = "grey60") + geom_point(colour = "#236192", size = 2) +
  labs(x = "Day", y = "M10 centre (clock hour)") +
  theme_actiRhythm()
```

## When to reach for each

Start with the actogram and the static summary. If the band is straight and the
nonparametric numbers sit comfortably, a cosinor and the interdaily and intradaily
metrics are enough. When the band bends, fragments, or wanders, the static numbers become
a compromise and the dynamic tools earn their place. Reach for the spectrogram to watch
the period and its strength over time, `sleep.changepoints()` to date a regime change,
singular spectrum analysis or empirical mode decomposition to read the circadian
component apart from the trend and the noise, multiscale entropy and the multifractal
spectrum to measure how the structure itself changes, and curve registration for a phase
that does not blur when the days disagree. Each one answers a question that a single
number cannot.

## References
