---
title: "Period and spectral analysis"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
bibliography: ../REFERENCES.bib
csl: ../apa.csl
link-citations: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, dpi = 110)
```

```{r libs, message = FALSE, warning = FALSE}
library(actiRhythm)
library(ggplot2)
```

## The idea, and a rule to remember

The cosinor assumes the period; period estimation **finds** it. Given a recording,
these tools ask "what is the dominant cycle length?", not "how well does a
24-hour cosine fit?" That matters whenever the rhythm may be free-running (a tau
that is not exactly 24), as it is in jet lag, shift work, dim-light protocols, and
many clinical and animal recordings. The two workhorses are the **Lomb-Scargle**
periodogram, the least-squares spectral estimator for unevenly sampled series
[@lomb1976; @scargle1982], and the **chi-square (Sokolove-Bushell)** periodogram,
the analysis-of-variance method long standard in chronobiology [@sokolove1978].

The rule to carry through this article: **the chi-square periodogram needs
near-regular sampling, while Lomb-Scargle tolerates gaps.** Actigraphy is rarely
gap-free (non-wear, dropped epochs, mixed epoch lengths), and a single device-off
stretch can pull the chi-square peak off the true period while Lomb-Scargle, which
works from the actual timestamps, stays on it. When in doubt with gappy data, trust
Lomb-Scargle.

## The math

Write the activity as values \(y_i\) (mean-subtracted) sampled at times \(t_i\),
\(i = 1, \dots, n\), and let \(\omega = 2\pi / \tau\) be the angular frequency of a
trial period \(\tau\). The **Lomb-Scargle** power at \(\tau\) is

$$
P(\omega) = \frac{1}{\sum_i y_i^2}\left[
  \frac{\bigl(\sum_i y_i \cos\omega(t_i - \delta)\bigr)^2}{\sum_i \cos^2\omega(t_i - \delta)}
  + \frac{\bigl(\sum_i y_i \sin\omega(t_i - \delta)\bigr)^2}{\sum_i \sin^2\omega(t_i - \delta)}
\right],
$$

where the time offset \(\delta\) is fixed by
\(\tan(2\omega\delta) = \sum_i \sin 2\omega t_i \big/ \sum_i \cos 2\omega t_i\). This
offset is exactly what makes the estimate invariant to the irregular spacing of the
\(t_i\): Lomb-Scargle is the least-squares fit of a sinusoid at each trial period.
Standard-normalised this way, the power lies in \([0, 1]\), and its significance
is read from the @baluev2008 analytic false-alarm probability.

The **chi-square** periodogram instead folds the series at each integer trial period
\(P\) (in epochs) into \(P\) phase bins over \(K = \lfloor N/P \rfloor\) whole cycles
and measures how much the phase means \(\bar A_h\) separate from the grand mean
\(\bar A\) [@sokolove1978]:

$$
Q_P = \frac{K\,N \sum_{h=1}^{P}(\bar A_h - \bar A)^2}{\sum_{i=1}^{N}(A_i - \bar A)^2}.
$$

Under the null of no rhythm at \(P\), \(Q_P \sim \chi^2_{P-1}\), which supplies the
significance line. The folding is where the regular-grid assumption enters: it treats
the \(i\)-th retained sample as if it sat at phase \(i \bmod P\), so a gap that
shifts later samples off their true phase distorts every \(Q_P\).

## Assumptions, and when they break

- **A long enough span.** A period in the 18-30 h band cannot be resolved from less
  than two days; `circadian.period()` enforces a >= 2-day span and >= 10 valid
  points and returns `NA` rather than guessing.
- **Stationarity.** Both periodograms assume one period holds across the recording.
  When tau drifts (re-entrainment, fragmentation), a single global number hides it.
  Use `circadian.spectrogram()` to see the period over time.
- **Regular sampling, for the chi-square periodogram only.** Its phase-folding needs
  a near-uniform grid; gaps mis-align the fold. Lomb-Scargle carries the real times
  and is unbiased under irregular sampling, the centrepiece of this article.
- **A search window.** The peak is taken within `[from, to]` (default 18-30 h). A
  rhythm outside the window cannot be found; widen it for ultradian or infradian work.

## Recovering known truth

Before trusting these tools on real data, watch them recover a period we plant. We
build a ten-day minute-level recording from a cosine whose period is **25 hours, not
24** (a deliberately free-running tau), then add noise.

```{r known-truth}
ts <- seq(as.POSIXct("2024-01-01", tz = "UTC"), by = 60, length.out = 10 * 1440)
th <- as.numeric(difftime(ts, ts[1], units = "hours"))
set.seed(1)
tau_true <- 25
counts <- pmax(0, 100 + 80 * cos(2 * pi * (th - 8) / tau_true) + rnorm(length(th), 0, 15))

ls0  <- circadian.period(counts, ts)
chi0 <- chi.sq.periodogram(counts, ts)
knitr::kable(
  data.frame(method   = c("Lomb-Scargle", "chi-square"),
             planted  = c(tau_true, tau_true),
             tau_h    = c(ls0$tau, chi0$period)),
  digits = 2, caption = "Both periodograms recover the planted 25-hour period."
)
```

Both land on 25, not the conventional 24: the method reports the period we put in,
not the one we expected. The confidence interval should then **bracket** the planted
value. `period.ci()` attaches a circular moving-block bootstrap interval that
respects the autocorrelation of activity [@kunsch1989; @politis1992].

```{r known-ci}
ci <- period.ci(counts, ts, n_boot = 200, seed = 1)
ci
tau_true >= ci$ci_lower && tau_true <= ci$ci_upper   # the interval brackets the truth
```

## On a real recording

The bundled recording runs the same way. `circadian.period()` returns a list whose
`tau` is the estimated endogenous period and whose `p_value` is the Baluev
false-alarm probability.

```{r real}
agd <- agd.counts(read.agd(example_agd(1), verbose = FALSE))
ls  <- circadian.period(agd$axis1, agd$timestamp)
c(tau = ls$tau, peak_power = ls$peak_power, p_value = ls$p_value,
  n_used = ls$n_used, span_days = round(ls$span_days, 2))
```

```{r ls-plot, fig.cap = "The Lomb-Scargle periodogram of the bundled recording over the 18-30 h window. The orange line marks the dominant period, the grey dashed line is 24 hours, and the red dashed line is the 0.05 Baluev false-alarm threshold, which is flat across periods (unlike the chi-square critical curve below). The peak sits to the right of 24, a longer-than-day dominant cycle."}
plot_periodogram(agd$axis1, agd$timestamp)
```

The bootstrap interval shows how firmly that period is pinned down. On this recording
the replicate periods spread widely across the search window, so the interval spans
several hours: a single tau on a noisy free-running record carries real uncertainty.

```{r ci-plot, fig.cap = "Bootstrap replicate periods (histogram) with the point estimate (orange) and the 95% confidence-interval band. A broad spread means the dominant period is only weakly determined."}
plot_period_ci(agd$axis1, agd$timestamp, n_boot = 200, seed = 1)
```

## Reading the numbers

State each output in human terms:

- **tau** is the dominant period in hours. Near 24 is entrained to the day; clearly
  above or below 24 suggests a free-running or mis-entrained rhythm. Read it against
  the 24-hour line, not in isolation.
- **peak_power** (Lomb-Scargle) runs 0 to 1; higher is a sharper, more dominant
  cycle. **Qp_peak** (chi-square) is on a chi-square scale and is judged against its
  per-period critical line.
- **p_value** is the chance a peak this tall arose from noise. Lomb-Scargle uses the
  Baluev analytic false-alarm probability; the chi-square periodogram reports a
  family-wise (Sidak) p-value across the scanned periods, so a single tall bar is not
  called real purely from multiple testing.
- **The confidence interval** is the honest precision of tau. A wide interval that
  straddles 24 is the data telling you it cannot separate an entrained from a
  free-running rhythm.

## When the sampling is gappy

This is the rule in action. We take the same planted-25 h series and **knock out a
contiguous block of epochs** (a roughly two-day device-off gap), then re-estimate.
Lomb-Scargle reads the real timestamps and should hold at 25; the chi-square fold,
which assumes a regular grid, should slip.

```{r wrong-way}
gap   <- th >= 5 * 24 & th < 7.3 * 24          # remove ~2.3 days in the middle
ts_g  <- ts[!gap]
cnt_g <- counts[!gap]

ls_g  <- circadian.period(cnt_g, ts_g)
chi_g <- chi.sq.periodogram(cnt_g, ts_g)
knitr::kable(
  data.frame(method = c("Lomb-Scargle", "chi-square"),
             truth  = c(tau_true, tau_true),
             full   = c(ls0$tau,  chi0$period),
             gappy  = c(ls_g$tau, chi_g$period)),
  digits = 2,
  caption = "After a multi-day gap, Lomb-Scargle stays on 25 h; the chi-square peak slips off."
)
```

The chi-square period moves off 25 (pulled toward a wrong value because the samples
after the gap no longer fold onto their true phase), while Lomb-Scargle, anchored to
the actual times, returns essentially the same tau as before. The lesson is the rule:
with gappy actigraphy, reach for Lomb-Scargle and treat a chi-square peak with
suspicion until you have confirmed the sampling is near-regular.

## The wider period family

The same spectral idea generates a family of tools that look past a single global
number.

**Period over time.** `circadian.spectrogram()` slides a window across the recording
and runs the chi-square periodogram in each, producing a period-by-time heat map that
shows drift, fragmentation, and re-entrainment a single fit averages away.

```{r spectrogram, fig.height = 4, fig.cap = "Sliding-window chi-square spectrogram of the bundled recording, coloured by Qp relative to its per-period significance threshold so equal colour means equal significance across periods. The white line traces the per-window peak period and the dashed line is 24 h."}
sg <- circadian.spectrogram(agd$axis1, agd$timestamp, step_hours = 12)
sg$plot
```

**Ultradian band power.** Rhythms faster than a day (the roughly 90-minute, 4-hour,
and 8-hour bands) carry real physiology. `ultradian.bandpower()` partitions the
activity variance into dyadic period bands with an undecimated Haar wavelet and
reports the fraction of variance in each [@percival2000].

```{r bandpower}
ultradian.bandpower(agd$axis1, agd$timestamp)
```

**What the cosinor leaves behind.** `residual.spectrum()` removes the fitted 24-hour
cosinor mean and estimates the spectrum of the residual, splitting it into ultradian
and high-frequency bands. Two recordings with the same 24-hour rhythm but different
residual fragmentation are told apart here.

```{r residual}
residual.spectrum(agd$axis1, agd$timestamp, period = 24)
```

## Limitations

- **Resolution scales with span.** The period grid is set by the recording length;
  short records give coarse, wide intervals. Distinguishing 24.0 from 24.5 h needs
  many days.
- **The chi-square periodogram is fragile to gaps.** As shown above, treat its peak
  with care on irregular sampling; prefer Lomb-Scargle, or gate and report wear time
  first.
- **Stationarity is assumed.** A single tau hides a drifting period; use the
  spectrogram when re-entrainment or fragmentation is plausible.
- **The window bounds the answer.** A rhythm outside `[from, to]` is invisible; set
  the window to the science, not the default.

## Reference and validation

The Lomb-Scargle estimator follows @lomb1976 and @scargle1982,
with the analytic false-alarm probability of @baluev2008 and the actigraphy
application of @ruf1999; the chi-square periodogram is that of @sokolove1978. The
period confidence interval uses the moving-block bootstrap of @kunsch1989 in its
circular form [@politis1992]. actiRhythm's Lomb-Scargle power is cross-checked
against a direct least-squares periodogram (the exact estimator: it uses the
double-angle time offset of @scargle1982, where the `lomb` package uses a
single-angle offset), and the chi-square \(Q_P\) statistic is validated against its
chi-square null, in the package's test suite.

## References
