---
title: "Phase, regularity, and social jet lag"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
bibliography: ../REFERENCES.bib
csl: ../apa.csl
link-citations: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, dpi = 110)
```

```{r libs, message = FALSE, warning = FALSE}
library(actiRhythm)
library(ggplot2)
```

## The idea, and a rule to remember

The metrics in this article are about **when** the rhythm happens and **how
consistently** it repeats, rather than how strong or how fragmented it is. The
Sleep Regularity Index asks whether you are asleep and awake at the same clock
times from one day to the next; activity onset and offset, the composite phase
deviation, and the phase-concentration tests pin down the timing of a daily
marker and how tightly it clusters; social jet lag measures the gap between your
work-day and free-day timing: the clock you keep for work versus the one your
body would choose.

The rule to carry through this article: **regularity is consistency of timing,
not quality of sleep.** The Sleep Regularity Index rewards a pattern that repeats
at the same clock time, so a person who is reliably awake all night every night
scores as *perfectly regular*. Read these timing metrics alongside a measure of
how much sleep there is, never as a stand-in for it.

## The math

Write the recording as a binary sleep/wake matrix \(s_{i,j}\), where \(i\) indexes
the \(M\) epochs of a day (clock-time bins) and \(j\) indexes the \(N\) calendar
days. The **Sleep Regularity Index** is the chance two consecutive days hold the
same state at the same clock time, rescaled to \([-100, 100]\) [@phillips2017]:

$$
\mathrm{SRI} = -100 + \frac{200}{M(N-1)}
  \sum_{i=1}^{M}\sum_{j=1}^{N-1} \mathbf{1}\!\left\{ s_{i,j} = s_{i,j+1} \right\}.
$$

Perfect repetition scores \(+100\), independent (coin-flip) days score about \(0\),
and a perfectly inverted pattern scores \(-100\).

**Social jet lag** compares the mid-point of sleep on free days (MSF) with that on
work days (MSW) [@wittmann2006]:

$$
\mathrm{SJL} = \mathrm{MSF} - \mathrm{MSW}.
$$

Because free-day sleep is often longer (catch-up sleep), the chronotype is read
from the **sleep-debt-corrected** mid-sleep MSFsc, which shifts the free-day
mid-point back by half the excess of free-day sleep over the weekly average
[@roenneberg2012; @roenneberg2003]:

$$
\mathrm{MSF_{sc}} = \mathrm{MSF} - \tfrac{1}{2}\!\left(\mathrm{SD_f} - \mathrm{SD_{week}}\right)
\quad\text{when } \mathrm{SD_f} > \mathrm{SD_w}.
$$

For the **phase-concentration** tests, write each day's marker as an angle
\(\theta_j = 2\pi t_j / 24\). The mean resultant length
\(R = \lVert N^{-1}\sum_j e^{i\theta_j} \rVert\) measures clustering, and the
Rayleigh statistic \(Z = N R^2\) tests it against uniformity [@fisher1993]. The
**composite phase deviation** combines each day's precision (deviation from the
person's own mean phase) and accuracy (deviation from a reference) as
\(\mathrm{CPD} = \overline{\sqrt{\text{precision}^2 + \text{accuracy}^2}}\)
[@fischer2016].

## Assumptions, and when they break

- **A scored sleep/wake state.** The SRI needs a per-epoch label, not raw counts.
  Build it from an activity scorer (`sleep.cole.kripke()` below); a different
  scorer or threshold can move the index.
- **Clock-time alignment, not elapsed time.** Epochs are binned by real clock
  time, so a recording that does not start at midnight is handled correctly, but
  a device whose clock is wrong will mistime every day.
- **Enough whole days.** The SRI is a between-consecutive-day quantity and the
  concentration tests are circular statistics; one or two days give an unstable
  answer. `sri.matrix()` reports how many consecutive-day pairs it actually used.
- **Missing data lowers power, not bias.** Non-wear epochs scored `NA` are skipped
  from the concordance count rather than read as sleep, but a recording that is
  mostly gaps leaves little to compare.

## Recovering known truth

Before trusting the SRI on real data, watch it behave on data whose answer we
know. We build three seven-day recordings and score each with the same Cole-Kripke
scorer: one with the **same** 23:00-07:00 sleep window every night, one whose
sleep onset **wanders** a few hours from day to day, and one where every epoch is
**independently** quiet or active. The SRI should fall from near \(100\) toward
\(0\) across the three.

```{r known-truth}
ts  <- seq(as.POSIXct("2024-01-01 00:00:00", tz = "UTC"), by = 60, length.out = 7 * 1440)
tod <- as.POSIXlt(ts)$hour + as.POSIXlt(ts)$min / 60
day <- as.integer(as.Date(ts) - as.Date(ts[1]))
set.seed(1)

# regular: the same nightly window; jittered: onset wanders; random: coin-flip epochs
regular_counts  <- ifelse(tod >= 23 | tod < 7, rpois(length(ts), 2), rpois(length(ts), 250))
onset           <- (23 + round(rnorm(7, 0, 3)))[day + 1]
jittered_counts <- ifelse((tod - onset) %% 24 < 8, rpois(length(ts), 2), rpois(length(ts), 250))
random_counts   <- ifelse(rbinom(length(ts), 1, 0.5) == 1, rpois(length(ts), 2), rpois(length(ts), 250))

sri_of <- function(counts) {
  state <- sleep.cole.kripke(counts)
  sleep.regularity.index(state, ts)
}
knitr::kable(
  data.frame(pattern = c("regular", "jittered", "random"),
             SRI = c(sri_of(regular_counts), sri_of(jittered_counts), sri_of(random_counts))),
  digits = 2,
  caption = "The SRI falls from its ceiling as a planted sleep pattern is made irregular, then random."
)
```

A sleep pattern that repeats at the same clock time returns the SRI at its ceiling
of \(100\); day-to-day jitter in the onset pulls it down sharply; coin-flip epochs
leave it near \(0\). The index reports exactly the regularity we put in.

## On a real recording

The recording bundled with the package runs the same way. We score it to a
per-epoch sleep/wake state, then read the SRI off the published epoch-by-day
matrix.

```{r real}
agd   <- agd.counts(read.agd(example_agd(1), verbose = FALSE))
state <- sleep.cole.kripke(agd$axis1)
sri.matrix(state, agd$timestamp)
```

`sleep.regularity.index()` is the same number with a lighter return: it calls
`sri.matrix()` internally, so the two always agree.

```{r real-sri}
sleep.regularity.index(state, agd$timestamp)
```

The phase markers run on the raw counts. `activity.onset.offset()` finds, on the
averaged day, the clock time where activity rises most sharply (onset) and falls
most sharply (offset) by the relative-difference method [@roenneberg2003].

```{r onset-offset}
activity.onset.offset(agd$axis1, agd$timestamp)
```

## Reading the numbers

- **SRI** runs \(-100\) to \(+100\). Above about \(80\) is a highly regular sleeper;
  the population mid-range sits near \(60\)-\(70\); below about \(40\) is markedly
  irregular. Here it is mid-range: the sleep/wake pattern repeats moderately well
  from day to day. Remember the rule: a high SRI says *consistent*, not *healthy*.
- **Activity onset / offset** are clock hours and are **circular**: an onset of
  23.8 and one of 0.2 are half an hour apart, not 23. Read them as the times the
  active day begins and ends.
- **Mid-sleep (MSW, MSF, MSFsc)** are clock times of the sleep mid-point; the
  corrected MSFsc is the chronotype you would compare across people.

## The wider phase-and-regularity family

The same timing idea generates a family of related descriptors.

**The gap between work and free days.** `social.jet.lag()` needs a small table of
sleep periods with `in_bed_time` and `out_bed_time`. We build a synthetic week:
five work nights to bed at 23:00 up at 07:00, two free nights to bed at 01:00 up
at 10:00, later and longer, as free days usually are.

```{r social-jet-lag}
dates <- as.Date("2024-01-01") + 0:6
free  <- weekdays(dates) %in% c("Saturday", "Sunday")
sleep_periods <- data.frame(
  in_bed_time  = as.POSIXct(ifelse(free, paste(dates + 1, "01:00:00"),
                                         paste(dates,     "23:00:00")), tz = "UTC"),
  out_bed_time = as.POSIXct(ifelse(free, paste(dates + 1, "10:00:00"),
                                         paste(dates + 1, "07:00:00")), tz = "UTC")
)
sjl <- social.jet.lag(sleep_periods, work_days = !free)
c(MSW = sjl$MSW_time, MSF = sjl$MSF_time, MSFsc = sjl$MSFsc_time,
  SJL_hours = sjl$social_jet_lag_hours, SJLsc_hours = sjl$social_jet_lag_sc_hours)
```

The free-day mid-sleep falls 2.5 hours later than the work-day one (the social
jet lag), and the sleep-debt correction trims that to the smaller SJLsc, the part
of the shift not explained by weekend catch-up sleep. A value above one hour is
the threshold commonly flagged as a health concern [@wittmann2006].

**Do the daily markers cluster?** A regular phase lands at the same clock time each
day. We take the per-day activity onset and test whether those onsets are
concentrated rather than scattered around the clock.

```{r per-day-onsets}
d <- as.Date(agd$timestamp)
onsets <- vapply(split(seq_along(d), d), function(ix) {
  if (length(ix) < 600) return(NA_real_)         # skip part-days
  activity.onset.offset(agd$axis1[ix], agd$timestamp[ix])$onset_h
}, numeric(1))
onsets <- onsets[!is.na(onsets)]
round(onsets, 2)
```

`phase.concentration()` reports the mean resultant length and the Rayleigh and
Hermans-Rasson tests of uniformity [@fisher1993; @landler2019]; a small p-value is
evidence the onsets cluster at a preferred time.

```{r phase-concentration}
phase.concentration(onsets)
```

```{r onset-plot, fig.height = 3, fig.cap = "The activity onset for each day. Tightly stacked points are a concentrated phase; the mean resultant length and Rayleigh test quantify that clustering."}
df <- data.frame(date = as.Date(names(onsets)), onset = onsets)
ggplot(df, aes(date, onset)) +
  geom_hline(yintercept = mean(onsets), colour = "grey70", linetype = 2) +
  geom_line(colour = "grey60") + geom_point(colour = "#236192", size = 2) +
  labs(x = NULL, y = "Activity onset (h)") +
  theme_actiRhythm()
```

**How far the phase strays.** `composite.phase.deviation()` summarises the same
onsets as a single instability score: the precision (scatter around the person's
own mean phase) and accuracy (distance from a reference) combined day by day
[@fischer2016].

```{r cpd}
composite.phase.deviation(onsets)
```

**Uncertainty in the mean phase.** `circadian.onset.ci()` gives a bootstrap
confidence interval for the mean onset time, resampling the per-day onsets on the
circle so the interval is correct even near midnight.

```{r onset-ci}
circadian.onset.ci(onsets)
```

## Limitations

- **The SRI depends on the scorer.** It is computed on a sleep/wake label, so a
  different scoring algorithm or threshold shifts it; report which scorer you used.
- **Regularity is not adequacy.** A perfectly regular short or mistimed sleeper
  still scores high. Pair the SRI with a sleep-duration or amplitude metric.
- **Social jet lag needs both kinds of day.** With no free nights (or no work
  nights) MSF or MSW is undefined and the difference is `NA`; a single free night
  gives an unstable free-day mid-point.
- **Circular statistics need several days.** The concentration tests, the CPD, and
  the onset confidence interval all return `NA` below three days and are noisy with
  only a handful.

## Reference and validation

The Sleep Regularity Index follows @phillips2017 in its
full epoch-by-day concordance form; social jet lag and the sleep-debt correction
follow @wittmann2006 and @roenneberg2012 [-@roenneberg2003]. The phase markers use
the relative-difference onset of @roenneberg2003, the composite phase deviation
of @fischer2016, and the circular tests of @fisher1993 and @landler2019.
actiRhythm's SRI and phase metrics are cross-checked against their reference
definitions in the [Validation](validation.html) article and the package's test
suite.

## References
