---
title: "From raw acceleration"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
bibliography: ../REFERENCES.bib
csl: ../apa.csl
link-citations: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5, dpi = 110)
```

```{r libraries, message = FALSE, warning = FALSE}
library(actiRhythm)
library(ggplot2)
```

This article continues from the get-started vignette and works from raw
accelerometer data rather than activity counts. It shows how actiRhythm reads raw
ActiGraph, Axivity, and GENEActiv files, auto-calibrates them, derives the ENMO,
MAD, and z-angle metrics, runs the full circadian analysis and a posture-based
sleep detector directly from a file, and checks both the per-epoch metrics and the
sleep window against GGIR. The objects below are rebuilt here so the article stands
on its own.

```{r data-setup, message = FALSE, warning = FALSE}
agd <- agd.counts(read.agd(example_agd(1), verbose = FALSE))
raw <- example_raw(days = 2)        # synthetic 2-day recording (or a real file path)

# Real .gt3x used only for the GGIR cross-check at the end. It lives in data-raw/
# (kept in the source repo, excluded from the CRAN build), so the cross-check runs
# when the repo is present and is skipped otherwise.
gt3x_file <- local({
  cands <- c("../../data-raw/MOS2E39230594.gt3x", "data-raw/MOS2E39230594.gt3x")
  inp <- tryCatch(knitr::current_input(dir = TRUE), error = function(e) NULL)
  if (!is.null(inp))
    cands <- c(file.path(dirname(inp), "..", "..", "data-raw", "MOS2E39230594.gt3x"), cands)
  root  <- tryCatch(rprojroot::find_root(rprojroot::has_file("DESCRIPTION")),
                    error = function(e) NULL)
  if (!is.null(root)) cands <- c(cands, file.path(root, "data-raw", "MOS2E39230594.gt3x"))
  hit <- cands[file.exists(cands)]
  if (length(hit)) normalizePath(hit[1]) else NA_character_
})
```

## Raw metrics and posture

Everything so far runs on activity counts. Raw accelerometer files also record the
gravity component, and with it body posture, which counts cannot carry. actiRhythm reads
the three common formats: ActiGraph `.gt3x`, Axivity `.cwa`, and GENEActiv `.bin`.

`raw.metrics()` takes a file path. Raw acceleration is far too large to bundle, so for
a reproducible illustration `example_raw()` synthesises a recording; a real file is read
the same way, with `raw.metrics("recording.cwa")`.

```{r raw-metrics}
m <- raw.metrics(raw, epoch = 60)   # per-epoch ENMO (mg), MAD, and the z-angle
head(m)
```

ENMO is the activity signal, and every method above takes it directly: pass it to
`circadian.rhythm()`, or let `circadian.raw()` run the whole analysis from the file.

```{r circadian-raw}
cr <- circadian.rhythm(m$ENMO, m$time)
c(IS = cr$IS, IV = cr$IV, RA = cr$RA)
```

Calibration is applied first inside `raw.metrics()`. `auto.calibrate()` finds the
per-axis gain and offset that return still periods to the 1 g sphere [@vanhees2014];
on data with a known distortion it returns the original gain and offset to within a
fraction of a percent.

```{r calibrate}
set.seed(1)
u <- matrix(rnorm(40 * 3), 40, 3); u <- u / sqrt(rowSums(u^2))
cal <- do.call(rbind, lapply(seq_len(40), function(i)
  matrix(rep(u[i, ] / c(1.03, 0.97, 1.01) + c(0.04, -0.03, 0.02), each = 300),
         300, 3) + rnorm(900, 0, 0.004)))
auto.calibrate(data.frame(x = cal[, 1], y = cal[, 2], z = cal[, 3]),
               fs = 30)[c("scale", "offset")]
```

The z-angle supports a sleep detector that needs no diary and works from posture,
which the counts cannot do. `rest.spt()` finds the nightly sleep-period-time window from the
distribution of angle change [HDCZA; @vanhees2018], `sib.vanhees()` scores
sustained-inactivity bouts [@vanhees2015], and `sleep.from.spt()` combines them into
onset, wake and efficiency. `detect.nonwear.raw()` flags a stationary, taken-off
device [low standard deviation and range over the hour; @vanhees2011], and passing
its mask as `wear` keeps a device-off stretch from being read as one long night.
Computed here on the synthetic recording at a 5-second epoch, with the detected
windows shaded.

```{r spt, fig.height = 3, fig.cap = "The z-angle of the synthetic recording at 5-second epochs, with the two nightly sleep-period-time windows detected from the angle alone (HDCZA), gated by raw non-wear, shaded."}
m5    <- raw.metrics(raw, epoch = 5, metrics = "anglez")
wear  <- detect.nonwear.raw(raw, epoch = 5)
spt   <- rest.spt(m5$anglez, m5$time, epoch_length = 5, wear = wear)
sib   <- sib.vanhees(m5$anglez, epoch_length = 5)
sleep <- sleep.from.spt(spt, sib, m5$time, epoch_length = 5)
thin <- seq(1, nrow(m5), by = 12)
ggplot() +
  geom_rect(data = spt, aes(xmin = onset, xmax = offset, ymin = -90, ymax = 90),
            fill = "#236192", alpha = 0.18) +
  geom_line(data = m5[thin, ], aes(time, anglez), linewidth = 0.2, colour = "grey30") +
  labs(x = NULL, y = "z-angle (deg)") +
  theme_actiRhythm()
sleep[, c("date", "onset", "offset", "tst", "efficiency")]
```

Two more estimators measure fragmentation, and both run on the counts already loaded:
`activity.balance.index()` maps a detrended-fluctuation exponent to a 0 to 1 score that
peaks at the healthy 1/f balance [@danilevicz2024], and `transition.probability()` gives
the closed-form rest-to-active and active-to-rest transition probabilities.

```{r abi-tp}
activity.balance.index(fractal.dfa(agd$axis1))
transition.probability(agd$axis1)[c("tp_ra_mle", "tp_ar_mle")]
```

### Agreement with GGIR

The raw metrics and the z-angle reimplement the van Hees algorithms that GGIR
established [@migueles2019], so they should reproduce GGIR's output, not merely be
internally consistent. The check below runs that comparison directly: it computes
GGIR's per-epoch ENMO and z-angle (`GGIR::g.getmeta()`) and actiRhythm's
(`raw.metrics()`) on the same real wrist recording and measures how closely they
agree.

That recording lives in the package's `data-raw/` folder (kept in the source
repository but excluded from the CRAN build, since a multi-day raw file is far too
large to ship), so the table is computed whenever the repository and GGIR are present
and skipped otherwise (`gt3x_file` is resolved in the setup chunk).

```{r ggir-agreement, eval = !is.na(gt3x_file) && requireNamespace("GGIR", quietly = TRUE) && requireNamespace("read.gt3x", quietly = TRUE), message = FALSE, warning = FALSE}
P <- GGIR::load_params()
P$params_general$windowsizes <- c(5, 900, 3600)
I <- GGIR::g.inspectfile(gt3x_file, params_rawdata = P$params_rawdata,
                         params_general = P$params_general)
ggir <- GGIR::g.getmeta(gt3x_file, params_rawdata = P$params_rawdata,
                        params_general = P$params_general,
                        params_cleaning = P$params_cleaning,
                        inspectfileobject = I)$metashort
acti <- raw.metrics(gt3x_file, epoch = 5)

# Align on the 5-second wall-clock label, then compare per epoch.
gg <- data.frame(k = gsub("T", " ", substr(as.character(ggir$timestamp), 1, 19)),
                 enmo = ggir$ENMO * 1000, anglez = ggir$anglez)
aa <- data.frame(k = format(acti$time, "%Y-%m-%d %H:%M:%S"),
                 enmo = acti$ENMO, anglez = acti$anglez)
m  <- merge(gg, aa, by = "k", suffixes = c("_ggir", "_acti"))

knitr::kable(
  data.frame(
    metric = c("ENMO (mg)", "z-angle (deg)"),
    cor    = c(cor(m$enmo_ggir, m$enmo_acti), cor(m$anglez_ggir, m$anglez_acti)),
    mad    = c(mean(abs(m$enmo_ggir - m$enmo_acti)),
               mean(abs(m$anglez_ggir - m$anglez_acti)))
  ),
  digits = c(0, 4, 2),
  col.names = c("Metric", "Correlation with GGIR", "Mean abs. difference"),
  caption = sprintf("Per-epoch agreement with GGIR over %s matched 5-second epochs.",
                    format(nrow(m), big.mark = ","))
)
```

Where the recording is present, the table shows the two implementations tracking each
other almost exactly over the whole file; the small residual is most likely epoch-edge
rounding rather than a difference in the algorithm. actiRhythm's van Hees
auto-calibration brings the recording onto the 1 g sphere first (the calibration demo
above recovers a planted gain and offset), which is what lets the ENMO values line up.

The same agreement holds for the **sleep-period (SPT) window**. Feeding the *same*
z-angle to actiRhythm's `rest.spt()` and to GGIR's HDCZA detector (`GGIR:::HASPT`)
isolates the detection algorithm from the upstream metric the table above already
validated, so any difference here would be in the algorithm itself.

```{r ggir-spt, eval = !is.na(gt3x_file) && requireNamespace("GGIR", quietly = TRUE) && requireNamespace("read.gt3x", quietly = TRUE), message = FALSE, warning = FALSE}
ps  <- GGIR::load_params()$params_sleep
spt <- rest.spt(acti$anglez, acti$time, epoch_length = 5)   # acti is from the chunk above
day <- as.Date(acti$time - 12 * 3600)                       # noon-to-noon nights

ggir_spt <- do.call(rbind, lapply(unique(day), function(dd) {
  idx <- which(day == dd)
  if (length(idx) < 720) return(NULL)
  h <- GGIR:::HASPT(acti$anglez[idx], params_sleep = ps, ws3 = 5,
                    HASPT.algo = "HDCZA", invalid = rep(0L, length(idx)))
  if (length(h$SPTE_start) == 0 || is.na(h$SPTE_start)) return(NULL)
  data.frame(date = as.Date(dd, origin = "1970-01-01"),
             ggir_onset  = acti$time[idx[round(h$SPTE_start)]],
             ggir_offset = acti$time[idx[round(h$SPTE_end)]])
}))

cmp <- merge(spt[, c("date", "onset", "offset", "duration")], ggir_spt, by = "date")
cmp$onset_diff_s  <- round(as.numeric(difftime(cmp$onset,  cmp$ggir_onset,  units = "secs")))
cmp$offset_diff_s <- round(as.numeric(difftime(cmp$offset, cmp$ggir_offset, units = "secs")))
knitr::kable(
  cmp[, c("date", "duration", "onset_diff_s", "offset_diff_s")],
  digits = 2,
  col.names = c("Night", "SPT duration (h)", "Onset diff (s)", "Offset diff (s)"),
  caption = "SPT window: actiRhythm rest.spt() vs GGIR HASPT (HDCZA) on the same z-angle."
)
```

Where the recording is present, the table shows the two implementations placing the
window within a single 5-second epoch on every night. The long-duration nights are
the parked-device tail (the watch lying still after it came off); a wear-time screen
removes them, but the two detectors still agree on what the raw algorithm returns
there, which is the point of the check. The sleep parameters `sleep.from.spt()`
derives (total sleep time, WASO, efficiency) are built on this window, so they
inherit the same agreement.

ActiGraph idle-sleep gaps are imputed as zeros by `read.gt3x`, which would otherwise
collapse the z-angle to a constant during quiescent periods. Like GGIR, actiRhythm
carries the last gravity vector forward through those gaps, so the angle stays correct
through the still periods the sleep detector relies on.

## References
