---
title: "Sleep and rest detection"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
bibliography: ../REFERENCES.bib
csl: ../apa.csl
link-citations: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, dpi = 110)
```

```{r libs, message = FALSE, warning = FALSE}
library(actiRhythm)
library(ggplot2)
```

## The idea, and a rule to remember

Sleep and rest detection turns a stream of activity counts into an answer to two
different questions. The **epoch scorers** (Cole-Kripke and Sadeh) label every
single minute as sleep or wake, giving a per-epoch state you can total into sleep
time. The **bout detectors** (change-point, Roenneberg, Crespo, and the
state-space model) step back and mark the *spans* of rest, the onset and offset
clock times of each consolidated period. One tells you *how much* sleep, the other
*when* it happened.

The rule to carry through this article: **decide first whether you want one main
sleep window per night or every rest bout the recording contains, because that
choice, not the algorithm's accuracy, is what separates these functions.**
`sleep.changepoints()` finds the single dominant rest bout of each circadian
cycle; `rest.periods()` and `rest.crespo()` return *every* consolidated bout, naps
included. Asking a one-bout method for naps, or a many-bout method for "the night",
is the most common way to misread the output.

## The math

The two epoch scorers are weighted windows over the counts. **Cole-Kripke**
[@cole1992] forms a sleep index from a seven-epoch window: four epochs before the
current one (\(P_4 \dots P_1\)), the current epoch \(C\), and two after
(\(N_1, N_2\)). Counts are scaled by 100 and capped:
$$
D = 0.001\,(106\,P_4 + 54\,P_3 + 58\,P_2 + 76\,P_1 + 230\,C + 74\,N_1 + 67\,N_2),
$$
and scores the epoch **sleep when \(D < 1\)**. Webster's rescoring rules
[@webster1982] then re-label short sleep bouts bracketed by sustained wake.

**Sadeh** [@sadeh1994] uses an eleven-epoch window (five each side) with counts
capped at 300:
$$
SI = 7.601 - 0.065\,\overline{A} - 1.08\,\mathrm{NAT} - 0.056\,s - 0.703\,\log(C+1),
$$
where \(\overline{A}\) is the window mean, \(\mathrm{NAT}\) the count of epochs in
\([50,100)\), \(s\) the standard deviation over the current and five preceding
epochs, and \(C\) the current count. The epoch is scored **sleep when
\(SI > -4\)**.

The change-point detector [@chensun2024] is geometric rather than weighted: it
fits a 24-hour cosinor to bound each rest and active span roughly, then places the
exact transition inside each bound at the split \(k\) minimising the within-segment
residual sum of squares,
$$
k^\* = \arg\min_k \;\sum_{i \le k}(x_i - \bar x_{\le k})^2 + \sum_{i > k}(x_i - \bar x_{>k})^2 .
$$

## Assumptions, and when they break

- **One-minute epochs.** Cole-Kripke and Sadeh were validated on 60-second epochs
  and their coefficients are tied to that resolution; both warn if `epoch_length`
  differs. Re-bin to one minute before scoring.
- **Counts, not raw acceleration.** The scorers expect ActiGraph-style activity
  counts on the vertical axis; feed them counts (`agd$axis1`), not raw g.
- **Non-wear is not sleep.** A taken-off device reads as deep rest. The scorers'
  `na_action` argument governs this: the default `"na"` emits `NA` for
  missing-count epochs so a device-off gap is not silently scored as sleep.
- **A clear day-night contrast for the bout detectors.** `sleep.changepoints()`,
  `rest.periods()`, and `rest.crespo()` all need enough rhythm to separate rest
  from activity; on a flat or very short record they return an empty result rather
  than erroring.

## Recovering known truth

Before trusting any of this on real data, plant an answer and check it comes back.
We build a seven-day recording with a **known nightly sleep window from 23:00 to
07:00** (near-zero activity at night, a busy day) and ask whether the epoch
scorers label the window sleep and whether `sleep.changepoints()` recovers the
23:00 onset and 07:00 wake.

```{r known-truth}
ts  <- seq(as.POSIXct("2024-01-01", tz = "UTC"), by = 60, length.out = 7 * 1440)
hod <- as.numeric(format(ts, "%H")) + as.numeric(format(ts, "%M")) / 60
set.seed(1)
asleep <- hod >= 23 | hod < 7                       # the planted sleep window
counts <- ifelse(asleep, 0, 250) + pmax(0, rnorm(length(ts), 0, 8))

ck <- sleep.cole.kripke(counts)
sd <- sleep.sadeh(counts)
knitr::kable(
  data.frame(
    scorer        = c("Cole-Kripke", "Sadeh"),
    sleep_recall  = c(mean(ck[asleep] == "S"),  mean(sd[asleep] == "S")),
    wake_recall   = c(mean(ck[!asleep] == "W"), mean(sd[!asleep] == "W"))),
  digits = 3,
  caption = "Both scorers label the planted 23:00-07:00 window sleep and the day wake."
)
```

Both scorers tag essentially all of the night as sleep and essentially all of the
day as wake. Now the change-point detector, which should place onset near 23:00 and
wake near 07:00 every night:

```{r known-changepoints}
cp <- sleep.changepoints(counts, ts)
cps_h <- as.numeric(format(cp$changepoints$time, "%H")) +
         as.numeric(format(cp$changepoints$time, "%M")) / 60
knitr::kable(
  data.frame(
    transition = c("sleep onset", "wake onset"),
    planted    = c(23, 7),
    recovered  = c(mean(cps_h[cp$changepoints$type == "sleep onset"]),
                   mean(cps_h[cp$changepoints$type == "wake onset"]))),
  digits = 2,
  caption = "Mean recovered transition time across the seven nights vs the planted values."
)
c(n_episodes = cp$n_episodes, mean_sleep_duration_h = cp$mean_sleep_duration)
```

The recovered onset and wake land within a minute of the planted 23:00 and 07:00,
the detector finds exactly seven nightly episodes, and the mean duration is the
eight hours we built in. The method recovers the truth we planted.

## On a real recording

The bundled recording runs the same calls. The two scorers return a per-epoch
state vector; `sleep.changepoints()` returns a per-night episode table.

```{r real}
agd <- agd.counts(read.agd(example_agd(1), verbose = FALSE))
table(cole_kripke = sleep.cole.kripke(agd$axis1))
table(sadeh       = sleep.sadeh(agd$axis1))
cp_real <- sleep.changepoints(agd$axis1, agd$timestamp)
cp_real
```

The two scorers agree closely on the total sleep fraction, and the change-point
detector resolves one main sleep episode per night with onset and wake clock times.

## Reading the numbers

- **`sleep_state` ("S"/"W")** from the scorers is a per-epoch label. Its useful
  summaries are the *count* of sleep epochs (multiply by the epoch length for sleep
  time) and the *runs* of consecutive sleep (sleep periods and their lengths).
- **Onset and wake times** from `sleep.changepoints()` are clock times; read them
  as "when did the main rest bout begin and end", and read `mean_sleep_duration` as
  the average nightly rest length.
- **Cole-Kripke vs Sadeh.** Cole-Kripke was validated for adults [@cole1992],
  Sadeh for children and adolescents [@sadeh1994]; pick by population. They will
  not agree epoch-for-epoch, and the disagreement is largest at the sleep-wake
  boundary, not in the middle of the night.

## One bout per night, or every bout? The rest detectors compared

This is the rule in action, and it stands in for the usual "wrong-way" demo: there
is no wrong algorithm here, only a wrong *question*. The same recording, run through
four detectors, gives four different bout counts, because they answer different
questions, not because three of them are wrong.

```{r contrast}
cp <- sleep.changepoints(agd$axis1, agd$timestamp)   # one main bout per night
rp <- rest.periods(agd$axis1, agd$timestamp)         # every bout (Roenneberg/MASDA)
rc <- rest.crespo(agd$axis1, agd$timestamp)          # every bout (Crespo morphology)
hm <- rest.hmm(agd$axis1, agd$timestamp, seed = 1)   # state-space alternative

knitr::kable(
  data.frame(
    detector = c("sleep.changepoints", "rest.periods", "rest.crespo", "rest.hmm"),
    question = c("one main bout / night", "every bout (naps incl.)",
                 "every bout (naps incl.)", "latent rest state"),
    bouts    = c(cp$n_episodes, rp$n_bouts, rc$n_rest_periods, NA),
    family   = c("CircaCP", "Roenneberg", "Crespo", "Gaussian HMM")),
  caption = "Four detectors, four counts: the difference is the question, not the accuracy."
)
```

`sleep.changepoints()` [@chensun2024] returns one episode per circadian cycle: the
single dominant night-time rest bout. `rest.periods()` [@roenneberg2015; @loock2021]
and `rest.crespo()` [@crespo2012] instead consolidate **every** rest bout, so a
daytime nap or a fragmented night becomes its own row. They reach that result by
two independent routes, Roenneberg's trend-and-correlation consolidation versus
Crespo's median-filter-and-morphology pipeline, which is why their bout counts
differ even though both aim to catch all rest. `rest.hmm()` [@huang2018hmm] is the
state-space alternative: rather than thresholding each epoch, it fits a Gaussian
hidden Markov model whose latent rest state *persists*, and reports a 24-hour
rest-occupancy profile.

```{r hmm-profile, fig.height = 3.4, fig.cap = "The HMM's probability of being in the rest state across the day. The state-space model infers the rest band, centred on the night, without a fixed count threshold."}
ggplot(hm$tod_profile, aes(hour, p_rest)) +
  geom_col(fill = "#236192") +
  scale_x_continuous(breaks = seq(0, 24, 6)) +
  labs(x = "Hour of day", y = "P(rest state)") +
  theme_actiRhythm()
```

The lesson is the rule: if you report `sleep.changepoints()` you are reporting *the
night*; if you report `rest.periods()` or `rest.crespo()` you are reporting *all
rest*, naps and all. Choose the one whose question matches yours, rather than
treating the differing bout counts as a contradiction.

## The wider sleep-and-rest family

**The state-space view in detail.** Beyond the occupancy profile above, `rest.hmm()`
returns the per-state emission means, the transition matrix, and a decoded
`state_path` with its own `sleep_state` vector, plus AIC/BIC for choosing between a
two- and three-state model [@huang2018hmm].

```{r hmm-detail}
hm$emission[, c("state", "label", "mean_transformed")]
c(time_at_rest = mean(hm$state_path == 1L),
  rest_persistence = hm$transition[1, 1], AIC = hm$AIC)
```

**The ultradian structure within sleep.** Once you have sleep windows,
`lids()` describes the sleep-cycle oscillation *inside* them. It needs a
`sleep_periods` data frame with `in_bed_time` and `out_bed_time` columns (which we
build directly from the change-point episodes), transforms activity to locomotor
inactivity (\(100/(\text{activity}+1)\)), and fits an ultradian cosine, reporting
the best period and its Munich Rhythmicity Index [@winnebeck2018].

```{r lids}
sleep_periods <- data.frame(
  in_bed_time  = cp$sleep_episodes$sleep_onset,
  out_bed_time = cp$sleep_episodes$wake_onset)
li <- lids(agd$axis1, agd$timestamp, sleep_periods)
li
```

The mean LIDS period lands in the ninety-minute-to-three-hour band expected of
human sleep cycles, recovered from the inactivity signal alone.

## Limitations

- **Epoch length is fixed for the scorers.** Cole-Kripke and Sadeh are only valid
  at one-minute epochs; do not port their coefficients to other resolutions.
- **Bout counts are not comparable across detectors.** A larger bout count from
  `rest.periods()` than from `sleep.changepoints()` is by design, not a sign one is
  better; compare a detector only against itself or against its own reference
  implementation.
- **No in-bed truth.** None of these methods see a diary or a true lights-off; they
  infer rest from movement, so a still-but-awake subject reads as rest and an active
  sleeper as wake. Where a diary exists, gate on it.
- **`lids()` depends on the supplied sleep windows.** Feeding it poorly placed
  windows propagates straight into the period estimate; build the windows from a
  detector you trust on the recording at hand.

## Reference and validation

The epoch scorers follow @cole1992 with the Webster rescoring
rules [@webster1982] and @sadeh1994. The bout detectors
implement the CircaCP change-point method of @chensun2024, the
Roenneberg / MASDA consolidation [@roenneberg2015] with the open re-implementation of
@loock2021, the @crespo2012 morphology pipeline, and the hidden-Markov rest-activity
model of @huang2018hmm; `lids()` follows @winnebeck2018.
actiRhythm's scorers and detectors are cross-checked against their reference
implementations (to the printed precision) in the [Validation](validation.html)
article and the package's test suite.

## References
