---
title: "Validation against reference packages"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
bibliography: ../REFERENCES.bib
csl: ../apa.csl
link-citations: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

actiRhythm computes its metrics from hand-written code rather than by wrapping
other packages, so it is fair to ask whether those metrics actually reproduce the
established implementations. This article answers that by running the reference
packages side by side on the recording that ships with actiRhythm, and
emitting the agreement from the code that follows: no number here is typed by
hand. Several of these checks also run in the package's own test suite (for example
`test-circadian-nparact.R` and `test-circadian-cosinor-extended.R`), so they are
exercised on every build.

Each comparison guards its reference package with `requireNamespace()`: a chunk
whose package is not installed shows its code but no output, so the article builds
either way. `ActCR` and the `cosinor` package are used live below; `nparACT` is
shown for you to run.

```{r data}
library(actiRhythm)

agd <- agd.counts(read.agd(example_agd(1), verbose = FALSE))
x   <- agd$axis1
ts  <- agd$timestamp

# The reference packages expect a whole number of 1440-minute days, so align to
# the first local midnight and keep only complete days.
mins  <- as.integer(format(ts, "%H")) * 60 + as.integer(format(ts, "%M"))
start <- which(mins == 0)[1]
x  <- x[start:length(x)]
keep <- (length(x) %/% 1440) * 1440
x  <- x[seq_len(keep)]
ts <- ts[start + seq_len(keep) - 1]

# actiRhythm's own metrics, and the averaged hourly profile the references reuse.
cr     <- circadian.rhythm(x, ts)
hourly <- t(apply(matrix(x, ncol = 1440, byrow = TRUE), 1,
                  function(d) tapply(d, rep(1:24, each = 60), mean)))
cat(sprintf("%d complete days of 60-second epochs\n", keep / 1440))
```

## Nonparametric metrics versus ActCR

`ActCR` is the Johns Hopkins reimplementation of the interdaily stability and
intradaily variability of @witting1990. Its `IS()` takes a
days-by-epochs matrix and `IV()` a chronological vector; on the same hourly profile
actiRhythm uses internally, the two implementations agree to the printed precision.

```{r actcr-isiv, eval = requireNamespace("ActCR", quietly = TRUE)}
cmp <- data.frame(
  metric     = c("IS", "IV"),
  actiRhythm = c(cr$IS, cr$IV),
  ActCR      = c(ActCR::IS(hourly), ActCR::IV(c(t(hourly))))
)
cmp$abs_diff <- abs(cmp$actiRhythm - cmp$ActCR)
knitr::kable(cmp, digits = 5, caption = "IS and IV: actiRhythm vs ActCR.")
```

## Nonparametric metrics versus nparACT

`nparACT` is the free reference implementation of the full nonparametric battery:
IS, IV, RA, L5, M10 [@witting1990; @vansomeren1999]. It reads a `time` /
`activity` data frame by name from the global environment and a sampling rate in
Hz. The chunk below reproduces the package's `test-circadian-nparact.R`: run it
with `nparACT` installed and it prints the agreement (IS/RA to two decimals, L5/M10
to within nparACT's rounding).

```{r nparact, eval = requireNamespace("nparACT", quietly = TRUE)}
assign("np_validation_df", data.frame(time = ts, activity = x), envir = globalenv())
np <- nparACT::nparACT_base("np_validation_df", SR = 1 / 60, plot = FALSE)
rm("np_validation_df", envir = globalenv())

cmp <- data.frame(
  metric     = c("IS", "IV", "RA", "L5", "M10"),
  actiRhythm = c(cr$IS, cr$IV, cr$RA, cr$L5, cr$M10),
  nparACT    = c(np$IS, np$IV, np$RA, np$L5, np$M10)
)
cmp$abs_diff <- abs(cmp$actiRhythm - cmp$nparACT)
knitr::kable(cmp, digits = 3, caption = "IS/IV/RA/L5/M10: actiRhythm vs nparACT.")
```

## Single cosinor versus the cosinor package

`cosinor.analysis()` fits the single-component cosinor of @cornelissen2014. The
`cosinor` package fits the same model through its formula
interface; given the same averaged daily profile, the MESOR and amplitude match
exactly.

```{r cosinor, eval = requireNamespace("cosinor", quietly = TRUE)}
ca   <- cosinor.analysis(x, ts)
prof <- data.frame(hour = rep(0:23, nrow(hourly)), Y = c(t(hourly)))
cl   <- cosinor::cosinor.lm(Y ~ time(hour) + 1, period = 24, data = prof)

cmp <- data.frame(
  parameter  = c("MESOR", "amplitude"),
  actiRhythm = c(ca$mesor, ca$amplitude),
  cosinor    = c(unname(cl$coefficients[1]), unname(cl$coefficients[2]))
)
cmp$abs_diff <- abs(cmp$actiRhythm - cmp$cosinor)
knitr::kable(cmp, digits = 5, caption = "Single cosinor: actiRhythm vs the cosinor package.")
```

## Extended (anti-logistic) cosinor versus ActCR

`cosinor.antilogistic()` fits the sigmoidally transformed cosine of @marler2006,
the same model as `ActCR::ActExtendCosinor()`. We validate
it on a controlled profile with a *moderate* transition, where every shape
parameter is identifiable. On a recording with near-square days, the steepness
`beta` is only weakly constrained once the rest-activity transition is sharp,
because the likelihood is then nearly flat in `beta`; the other parameters still
agree closely while `beta` itself is not pinned down, so we show the well-posed
case here.

```{r antilog, eval = requireNamespace("ActCR", quietly = TRUE)}
# A 3-day synthetic recording whose average day is a moderate sigmoid.
set.seed(1)
n   <- 1440 * 3
tse <- seq(as.POSIXct("2024-01-01", tz = "UTC"), by = 60, length.out = n)
hr  <- as.numeric(format(tse, "%H")) + as.numeric(format(tse, "%M")) / 60
ctp <- cos((hr - 15) * 2 * pi / 24)
ce  <- 50 + 250 * plogis(3 * (ctp - 0.1)) + rnorm(n, 0, 12)
ce[ce < 0] <- 0

# The 1440-minute averaged profile ActExtendCosinor consumes.
idx  <- as.integer(format(tse, "%H")) * 60 + as.integer(format(tse, "%M"))
prof1440 <- vapply(0:1439, function(i) mean(ce[idx == i], na.rm = TRUE), numeric(1))

fit <- cosinor.antilogistic(ce, tse)
ref <- ActCR::ActExtendCosinor(prof1440, window = 1)$params

cmp <- data.frame(
  parameter  = c("minimum", "amplitude", "alpha", "beta", "acrotime"),
  actiRhythm = c(fit$minimum, fit$amplitude, fit$alpha, fit$beta, fit$acrophase),
  ActCR      = c(ref$minimum, ref$amp, ref$alpha, ref$beta, ref$acrotime)
)
cmp$abs_diff <- abs(cmp$actiRhythm - cmp$ActCR)
knitr::kable(cmp, digits = 3, caption = "Anti-logistic cosinor: actiRhythm vs ActExtendCosinor.")
```

## Raw metrics versus GGIR

The raw-acceleration chain (van Hees auto-calibration, ENMO/MAD, and the
diary-free z-angle sleep detector) reimplements the algorithms `GGIR` established,
and is cross-checked against GGIR directly. On a real wrist recording kept in the
source repository (a multi-day raw file is too large for the CRAN build), the
[From raw acceleration](raw-pipeline.html) article computes per-epoch ENMO and the
z-angle against `GGIR::g.getmeta()`, and the sleep-period window against GGIR's
HDCZA detector (`HASPT`). Those tables are computed live whenever the repository and
GGIR are present; they report the per-epoch correlations and the window agreement
directly, and the package's opt-in `test-ggir.R` re-runs the same comparison and
fails if the agreement degrades.

## Bottom line

On the bundled recording, actiRhythm's interdaily stability, intradaily
variability, and single cosinor reproduce `ActCR` and the `cosinor` package to the
printed precision; the full nonparametric battery matches `nparACT` to its rounded
output; and the anti-logistic cosinor matches `ActExtendCosinor` on an
identifiable profile. These are not one-off checks: they are part of the
package's test suite and run on every build. The methods behind them are
@witting1990, @vansomeren1999, @cornelissen2014, @nelson1979, and @marler2006.

## References
